---
title: "Untitled"
output: html_document
date: "2026-02-10"
---






# Contigs-based analysis of CAZyme genes



# 1. Load data


```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)

setwd('C:\\Users\\lehakkin\\RStudio_computer')

load(file = 'CAZyme_ps_COVERAGE_FINAL')#ps
ps

meta <- meta(ps)
```



# 2. Heatmap of CAZymes



I will do heatmap for CAZyme family RA values.

For the final heatmap figure the RA values will be standardizes (z-transformed)

First pick the 20 most abundant CAZyme family in each soil type depth

Make a HEATMAP of the most abundant CAZyme family


## 2.1.  Pick 20 most abund CAZymes in each soil management layer

```{r}
ps_RA <- transform(ps, "compositional")
CAZy_family <- aggregate_rare(ps_RA, level = 'enzyme', detection = 0/100, prevalence = 0/140)
CAZy_family
```



```{r}
# sample wise filtering according to most abund. genera

# Initialize an empty list to store the taxa
abund.taxa <- list()

for (i in meta$sample_type) {
  for (j in meta$depth) {
    x <- sample_names(sample_data(CAZy_family)[sample_data(CAZy_family)$sample_type == i & sample_data(CAZy_family)$depth == j,])

# Calculate taxa mean of the selected samples
top20 <- head(sort(rowMeans(otu_table(CAZy_family)[,x]), decreasing = TRUE), 20)

result_name <- paste("sample_type", i, "depth", j, sep = "_")

abund.taxa[[result_name]] <- top20
  }
}

```



## 2.1. Combine picked taxa

```{r}
management_layer <- c(names(abund.taxa))

all_top20 <- c()

for (i in management_layer) {
  top20 <- c(names(abund.taxa[[i]]))
  all_top20 <- c(all_top20, top20)

}

all_top20_unique <- unique(all_top20)
length(all_top20_unique)

all_top20_unique_df <- as.data.frame(all_top20_unique)

#setwd('C:\\Users\\lehakkin\\RStudio_computer')
#write.csv2(all_top20_unique_df, file = "all_top20_CAZymes_unique.csv", row.names = FALSE)
```

After the filtering, there is 40 CAZyme families left out of the 295 in the beginning


## 2.3. Make a phyloseq of picked taxa


```{r}
#setwd('C:\\Users\\lehakkin\\RStudio_computer')
#all_top20_unique_df <- read.csv2(file = "all_top20_CAZymes_unique.csv")
all_top20_unique <- unique(all_top20_unique_df$all_top20_unique)
# filter by taxa name
CAZy_family_pruned <- prune_taxa(all_top20_unique, CAZy_family)
CAZy_family_pruned

# make a tax file with the group still available that I'll need later
tax_full <- as.data.frame(tax_table(ps_RA))
##remove duplicate rows across family
tax_full <- tax_full[!duplicated(tax_full[c('enzyme')]), ] 
tax_subset <- as.data.frame(tax_table(CAZy_family_pruned))
## combine these
tax <- left_join(tax_subset, tax_full, by="enzyme")

# remove extra
rm(tax_full)
rm(tax_subset)
```


## 2.4. MAKE HEATMAP figure

Note! I will not do Z-transforming the data before plotting the heatmap. Instead I will do row scaling in the Pheatmap stage, which means that each gene's relative abundance is standardized

```{r}
meta <- meta(CAZy_family_pruned)
meta$sample_type <- as.factor(meta$sample_type)
meta$sample_type <- factor(meta$sample_type, levels=c('forest', 'meadow', 'organic', 'conventional'))

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

# Z-transform
#CAZy_family_pruned_z <- microbiome::transform(CAZy_family_pruned, 'Z', log10 = FALSE)
CAZy_family_pruned_z <- CAZy_family_pruned

data_subset <- as(otu_table(CAZy_family_pruned_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 43)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]
```



```{r}
library("pheatmap")
library("ggplotify")

# Create color palette
my_colour <- list(
  Treatment = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional = "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA", '30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")
)

# Add gap between sample_types
gp_col <- which(diff(as.numeric(factor(my_sample_col3$Treatment))) != 0)

# Order the tax table - first by family, then by family (both alphabetically)
tax <- tax[order(tax$family, tax$enzyme), ]
rownames(tax) <- tax$enzyme

# Reorder f4 to match the sorted tax table
f4 <- f4[rownames(tax), ]

# Find where family changes occur (for gaps)
gp_row <- which(diff(as.numeric(factor(tax$family))) != 0)

# Create the heatmap plot
p2 <- as.ggplot(function() {
  pheatmap(
    f4,
    cluster_cols = FALSE,
    cluster_rows = FALSE,
    annotation_col = my_sample_col3,
    annotation_colors = my_colour,
    gaps_col = gp_col,
    gaps_row = gp_row,
    color = colorRampPalette(c("navy", "white", "red"))(50),
    show_colnames = FALSE,
    legend = TRUE,
    border_color = NA,
    cellheight = 16,
    fontsize = 14,
    fontsize_row = 14,
    annotation_names_row = FALSE,
    annotation_names_col = FALSE,
    scale = "row"
  )
})

print(p2)
```






# 3.Heatmap of CAZyme classes




```{r}
ps_RA <- transform(ps, "compositional")
CAZy_family <- aggregate_rare(ps_RA, level = 'family', detection = 0/100, prevalence = 0/140)
CAZy_family
```

Now no need to filter anything



Note! I will not do Z-transforming the data before plotting the heatmap. Instead I will do row scaling in the Pheatmap stage, which means that each gene's relative abundance is standardized

```{r}
meta <- meta(CAZy_family)
meta$sample_type <- as.factor(meta$sample_type)
meta$sample_type <- factor(meta$sample_type, levels=c('forest', 'meadow', 'organic', 'conventional'))

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

# Z-transform
#CAZy_family_pruned_z <- microbiome::transform(CAZy_family_pruned, 'Z', log10 = FALSE)
CAZy_family_pruned_z <- CAZy_family

data_subset <- as(otu_table(CAZy_family_pruned_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 11)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]
```



```{r}
library("pheatmap")
library("ggplotify")

# Create color palette
my_colour <- list(
  Treatment = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional = "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA", '30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")
)

# Add gap between sample_types
gp_col <- which(diff(as.numeric(factor(my_sample_col3$Treatment))) != 0)

# Order the tax table - first by family, then by family (both alphabetically)
tax <- as.data.frame(tax_table(CAZy_family))
#tax <- tax[order(tax$family, tax$family), ]
#rownames(tax) <- tax$family

# Reorder f4 to match the sorted tax table
f4 <- f4[rownames(tax), ]

# Find where family changes occur (for gaps)
gp_row <- which(diff(as.numeric(factor(tax$family))) != 0)

# Create the heatmap plot
p2 <- as.ggplot(function() {
  pheatmap(
    f4,
    cluster_cols = FALSE,
    cluster_rows = TRUE,
    annotation_col = my_sample_col3,
    annotation_colors = my_colour,
    gaps_col = gp_col,
    #gaps_row = gp_row,
    color = colorRampPalette(c("navy", "white", "red"))(50),
    show_colnames = FALSE,
    legend = TRUE,
    border_color = NA,
    cellheight = 16,
    fontsize = 14,
    fontsize_row = 14,
    annotation_names_row = FALSE,
    annotation_names_col = FALSE,
    scale = "row"
  )
})

print(p2)
```



# 4. PCoA for CAZymes


```{r}
ps_RA <- transform(ps, "compositional")
ps_x <- aggregate_rare(ps_RA, , level = 'enzyme',
    detection = 0/100, prevalence = 0/140)

ps_x
```



```{r}
otu = as(otu_table(ps_x), "matrix")
# transpose if necessary
if(taxa_are_rows(ps_x)){otu <- t(otu)}
# Coerce to data.frame
otu = as.data.frame(otu)
otu <- otu %>% replace(is.na(.), 0)
otu <- as.matrix(otu)

bray_dist <- vegan::vegdist(otu, method="bray")
str(bray_dist)

# use k = 3 so we'll get calculations for three axes
pcoa <- cmdscale(bray_dist, eig=TRUE, k = 3)
ordiplot(scores(pcoa),display="sites", type="points")
```


Let's first make PCoA ordination with axes 1 and 2


Env. variables

```{r}
# Post-Hoc Projections of environmental variables
# envfit in vegan projects points onto vectors that have maximum correlation with corresponding environmental variables
pcoa.env12 <- envfit(pcoa, meta[ ,c("pH_H2O", "EC_uScm", "C_g_per_kg", "N_gkg", "TP_gkg", "sample_type", "depth_numerical", "DOC_mgkg", "Pinorg_mgkg", "Porg_mgkg", "log_root", "C_per_N", "Feox_mmolkg", "Alox_mmolkg", "oxides_mmolkg", "PH2O_mgkg")], na.rm = TRUE, choices=c(1:2), permutations = 999)


# main effects
pcoa.env12

ef12.adj <- pcoa.env12
pvals.adj <- p.adjust (pcoa.env12$vectors$pvals, method = 'bonferroni')
ef12.adj$vectors$pvals <- pvals.adj
ef12.adj

```


To plot (classical) mds (which is equivalent to PCoA) with ggplot a new datasheet needs to be created which contains the x,y points for each site. You can do this by calling the scores of you mds.


```{r}
site.scrs <- as.data.frame(scores(pcoa, display = "sites")) #save pcoa results into dataframe
site.scrs <- cbind(site.scrs, soil_type = meta$sample_type) #add grouping variable "soil_type" to dataframe
site.scrs <- cbind(site.scrs, depth = meta$depth) #add grouping variable of depth

head(site.scrs)
```


I will be following somewhat this tutorial for fitting the environmental variables etc:

[jkzorz github](https://jkzorz.github.io/2020/04/04/NMDS-extras.html)

To show environmental extrinsic variables another datasheet needs to be created

Citation from the [jkzorz github](https://jkzorz.github.io/2020/04/04/NMDS-extras.html)

*"Extracting the required information from the envfit result is a bit more complicated. The envfit output contains information on the length of the segments for each variable. The segments are scaled to the r2 value, so that the environmental variables with a longer segment are more strongly correlated with the data than those with a shorter segment. You can extract this information with scores. Then these lengths are further scaled to fit the plot. This is done with a multiplier that is analysis specific, and can be accessed using the command ordiArrowMul(en). Below I multiply the scores by this multiplier to keep the coordinates in the correct proportion."*

Because my data contained continuous and categorical environmental variables, Im extracting the information from both separately using the vectors and factors options respectively.


Categorial and continuous variables

```{r}

# first categorial variables
env.scores_cat12 <- as.data.frame(scores(pcoa.env12, display = "factors"))*0.25 # multiply the envfit scores to keep them in the frame of the ordination
env.scores_cat12 <- cbind(env.scores_cat12, env.variables = rownames(env.scores_cat12)) #and then gives them their names

env.scores_cat12 <- cbind(env.scores_cat12, pval = pcoa.env12$factors$pvals) # add pvalues to dataframe
sig.env.scores_cat12 <- subset(env.scores_cat12, pval<=0.05) #subset data to show variables significant at 0.05
sig.env.scores_cat12
# all were significant


# then continous variables
env.scores_cont12 <- as.data.frame(scores(pcoa.env12, display = "vectors"))*0.25 # multiply the envfit scores to keep them in the frame of the ordination
env.scores_cont12 <- cbind(env.scores_cont12, env.variables = rownames(env.scores_cont12)) #and then gives them their names

env.scores_cont12 <- cbind(env.scores_cont12, pval = pcoa.env12$vectors$pvals) # add pvalues to dataframe
sig.env.scores_cont12 <- subset(env.scores_cont12, pval<=0.05) #subset data to show variables significant at 0.05
sig.env.scores_cont12
# all were significant
```


Now we have the relevant information for plotting the ordination in ggplot


```{r}
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")

site.scrs$new_depth <- site.scrs$depth

site.scrs$new_depth <- gsub("...", "-", site.scrs$new_depth, fixed = TRUE)

site.scrs$new_depth[site.scrs$new_depth=="40-"] <- "40-80"

site.scrs$new_depth[site.scrs$new_depth=="0-10"] <- "0-10 cm"
site.scrs$new_depth[site.scrs$new_depth=="10-20"] <- "10-20 cm"
site.scrs$new_depth[site.scrs$new_depth=="20-30"] <- "20-30 cm"
site.scrs$new_depth[site.scrs$new_depth=="30-40"] <- "30-40 cm"
site.scrs$new_depth[site.scrs$new_depth=="40-80"] <- "40-80 cm"
```


Get axis %

```{r}
#First lets run the PCoA with phyloseq to get the axis percentages (plot_ordination gives axis percentages)

GP.ord <- ordinate(ps_x, "PCoA", "bray")
p2 = plot_ordination(ps_x, GP.ord, type="samples", color="sample_type", shape="depth")
p2
```

Remember to change the axis percentages accordingly below!!!

Plot


```{r}
pcoa.plot <- ggplot() + geom_point(data=site.scrs, aes(Dim1, Dim2, colour = factor(site.scrs$soil_type), shape = factor(site.scrs$new_depth)), size = 6, alpha = 0.6, stroke = 1.5) + theme_cowplot() + theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid")) + labs(colour = "", shape = "") + theme(legend.text = element_text(size = 12), axis.text = element_text(size = 16)) + scale_colour_manual(values = MyPalette) + labs(y = "PCoA2 (14.5%)", x = "PCoA1 (29.6%)")


pcoa.plot

```

Add env. variables to the PCoA

First, simplify the names


```{r}
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "pH_H2O"] <- "pH"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "C_g_per_kg"] <- "C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "N_gkg"] <- "N"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "TP_gkg"] <- "P-tot"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "depth_numerical"] <- "Depth"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "DOC_mgkg"] <- "DOC"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Porg_mgkg"] <- "P-org"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "log_root"] <- "log root"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "C_per_N"] <- "C/N"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Feox_mmolkg"] <- "Fe-ox"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Alox_mmolkg"] <- "Al-ox"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "depth"] <- "Depth"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "log root"] <- "Root biomass"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "S_mg_per_kg"] <- "S"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Mn_mg_per_kg"] <- "Mn"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "PH2O_mgkg"] <- "P-H2O"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "POM_C_percent"] <- "POM-C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "MAOM_C_percent"] <- "MAOM-C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "oxides_mmolkg"] <- "Oxides"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "EC_uScm"] <- "EC"
```




...then plot

```{r fig.dim = c(12, 12)}
p1 <- pcoa.plot +
  geom_segment(aes(x = 0, y = 0, xend = Dim1, yend = Dim2), data = sig.env.scores_cont12, size =1, alpha = 0.3, colour = "grey30") + ggrepel::geom_text_repel(data = sig.env.scores_cont12, aes(x = Dim1, y = Dim2), colour = "blue", fontface = "bold", label = sig.env.scores_cont12$env.variables, segment.size = 0.2, box.padding = unit(0.1, "lines"), point.padding = (0.1), force = 1, max.time = 30, nudge_y = 0.00, nudge_x = 0.00) + theme(legend.position = "right", legend.text = element_text(size = 12))

p1
```






# 5. PERMANOVA

```{r}
ps_RA <- transform(ps, "compositional")
ps_x <- aggregate_rare(ps_RA, , level = 'enzyme',
    detection = 0/100, prevalence = 0/140)

ps_x
```


Calculate Bray-Curtis (dis)similarities

```{r}
ps_RA_bray <- phyloseq::distance(ps_x, method = "bray")
```


Homogenity of variances

Check that variance homogeneity assumptions hold (to ensure the reliability of the results). If  groups have signif. different spreads the permanova result may be potentially explained by that, rtaher than the groups.

Betadisper first calculates the average distance of group members to the group centroid in multivariate space (generated by a distance matrix). Then, an ANOVA is done to test if the dispersions (variances) of groups are different.


for management type

```{r}
anova(betadisper(ps_RA_bray, meta$sample_type))
```
         

We see that the ANOVA p-value is not significant meaning that the **homogeneity of variance assumption is met**


for depth

```{r}
anova(betadisper(ps_RA_bray, meta$depth))
```

We see that the ANOVA p-value is  significant meaning that **homogeneity of variance assumption is *NOT* met**

post hoc

I'll do post hoc analysis with Tukey's test to see which groups differ in relation to their variances

```{r}
TukeyHSD(betadisper(ps_RA_bray, meta$depth))
```

Dispersions differ significantly between the deepest layer and the ones above it


## 5.1. Initial PERMANOVAs


First, I will do PERMANOVA so that I include all management types (later without forest)


```{r}
# first with just soil type and strata option
adonis2(formula = ps_RA_bray~ sample_type, data = meta, permutations = 999, method = "bray", by = "margin", strata = meta$depth)
```

 treatment is significant


Depth

```{r}
# first with just soil type and strata option
adonis2(formula = ps_RA_bray~ depth, data = meta, permutations = 999, method = "bray", by = "margin", strata = meta$sample_type)
```

With P value = 1e-04, the soil layer is significant and R2 value indicated 29% variance explanation


## 5.2. Final PERMANOVA


```{r}
set.seed(1)
final <- adonis2(formula = ps_RA_bray ~depth * sample_type, data = meta, permutations = 9999, method = "bray", by = "terms")
final
```


# 6. Add PERMANOVA results to PCoA



```{r}
library(ggpubr) ## for getting stat_regline_equation
library("ggpmisc")

p <- p1 + 
  annotate(
    geom = "text",
    x = -0.25,
    y = 0.28,
    label = "PERMANOVA\nSoil layer:Treatment: R²=0.09; p=0.042 \nSoil layer: R²=0.29; p<0.001\nTreatment: R²=0.05; p<0.001\nModel R²=0.42",
    hjust = 0,
    size = 4  # Increase this value until text matches your desired size (default is 3.88)
  )

p
```



# 7. Differential abundance analysis: MaAsLin2


## 7.1. CAZyme enzymes


MaAsLin2 (Microbiome Multivariate Analysis)
Handles relative abundance data.

Works with generalized linear models (GLMs).


I will be using the top 20 treatment-soil-layer CAZymes only, use the coverage clr-transformed RA-values 

I will not be using the forest due to too few replicates





```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'enzyme', detection = 0/100, prevalence = 0/140)
ps_x

#setwd('C:\\Users\\lehakkin\\RStudio_computer')
#all_top20_unique_df <- read.csv2(file = "all_top20_CAZymes_unique.csv")
all_top20_unique <- unique(all_top20_unique_df$all_top20_unique)

ps_x <- prune_taxa(all_top20_unique, ps_x)
ps_x

tax <- as.data.frame(tax_table(ps_x))
```

### 7.1.1. With meadow

```{r}
#To remove those from your phyloseq object
ps_x <- subset_samples(ps_x, sample_type!="forest")
#This will return a ps object with the samples removed
ps_x

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```



```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CAZyme_enzyme')


# Make input_data
## input_data = dataframe where named gene_data where rows = samples and columns = genes
## input_metadata = dataframe here named metadata)


## get the otu table
gene_data <- as.data.frame(otu_table(ps_x))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_x)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "Treatment_Layer_WITH_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow')

```



### 7.1.2.Without meadow


```{r}
#To remove those from your phyloseq object
ps_x <- subset_samples(ps_x, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_x

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```



```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CAZyme_enzyme')


# Make input_data
## input_data = dataframe where named gene_data where rows = samples and columns = genes
## input_metadata = dataframe here named metadata)


## get the otu table
gene_data <- as.data.frame(otu_table(ps_x))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_x)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "Treatment_Layer_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional')

```



## 7.2. CAZyme classes


```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'family', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))
```


### 7.2.1. With meadow



```{r}
#To remove those from your phyloseq object
ps_x <- subset_samples(ps_x, sample_type!="forest")
#This will return a ps object with the samples removed
ps_x

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```


```{r}
# Run MaAsLin2
library(Maaslin2)

# Make input_data
## input_data = dataframe where named gene_data where rows = samples and columns = genes
## input_metadata = dataframe here named metadata)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CAZyme_Family')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_x))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_x)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "Treatment_Layer_With_Meadow",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow')

```




### 7.2.2. Without meadow



```{r}
#To remove those from your phyloseq object
ps_x <- subset_samples(ps_x, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_x

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

```


```{r}
# Run MaAsLin2
library(Maaslin2)

# Make input_data
## input_data = dataframe where named gene_data where rows = samples and columns = genes
## input_metadata = dataframe here named metadata)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CAZyme_Family')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_x))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_x)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "Treatment_Layer_WO_Meadow",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional')

```





# 8. Figures: boxplots





```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'family', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```

## 8.1. AA



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

AA <- ggplot(df, aes(x = Layer, y = AA, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "AA") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

AA

```



```{r}
names(gene_data)
```


## 8.2. CBM

```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

CBM <- ggplot(df, aes(x = Layer, y = CBM, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "CBM") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

CBM

```




```{r}
names(gene_data)
```


## 8.3. CE



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

CE <- ggplot(df, aes(x = Layer, y = CE, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "CE") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

CE

```



```{r}
names(gene_data)
```



## 8.4. cohesin


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

cohesin <- ggplot(df, aes(x = Layer, y = cohesin, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Cohesin") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

cohesin

```





```{r}
names(gene_data)
```



## 8.5. GH


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

GH <- ggplot(df, aes(x = Layer, y = GH, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "GH") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

GH

```







```{r}
names(gene_data)
```



## 8.6. GT




```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

GT <- ggplot(df, aes(x = Layer, y = GT, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "GT") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

GT

```




```{r}
names(gene_data)
```




## 8.7.PL




```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
PL <- merge(metadata, gene_data, by=0, all.x=TRUE)

PL <- ggplot(df, aes(x = Layer, y = PL, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "PL") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

PL

```





```{r}
names(gene_data)
```




## 8.8. SLH




```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

SLH <- ggplot(df, aes(x = Layer, y = SLH, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "SLH") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

SLH

```






```{r}
names(gene_data)
```






# 9. CAZyme PLant vs Microbial SUBSTRATES

```{r}
substrates <- read.csv2(file = "C:\\Users\\lehakkin\\RStudio_computer\\CAZyme_Substrates_from_Huan_et_al_2024.csv")
tax <- as.data.frame(tax_table(ps))
tax <- left_join(tax, substrates, by=c("enzyme"="CAZyme_family"))
tax <- as.data.frame(tax)
tax <- tax[, c(3:5, 1,2)]
row.names(tax) <- tax$enzyme

tax_table(ps) <- tax_table(as.matrix(tax))
ps
rank_names(ps)
```

## 9.1. Figures


Plant and Microbial derived C decomposing



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'Derived', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Microbial <- ggplot(df, aes(x = Layer, y = Microbial, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Microbial-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Microbial

```


```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Plant <- ggplot(df, aes(x = Layer, y = Plant, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Plant-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Plant

```



## 9.2. Test


```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'Derived', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```


Treatment in ecah soil layer separately


```{r}
ps_y <- subset_samples(ps_x, depth=="20...30")
ps_y
```


```{r}
library(car)

x_df <- psmelt(ps_y)

taxa <- unique(x_df$OTU)

for (i in taxa) {
 df <- filter(x_df, OTU == i)
 print(i)
 # Using leveneTest()
result = leveneTest(Abundance ~ sample_type, df)
# print the result
print(result)

}

x <- x_df %>%
  filter(OTU %in% taxa) %>% group_by(OTU, sample_type) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

```

All are homogenic


```{r}
y_df <- psmelt(ps_x)
x <- y_df %>%
  group_by(OTU, sample_type, depth) %>% summarise(mean = round(mean(Abundance, na.rm = TRUE)*100,1), se = round((sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))) * 100,1))
x <- subset(x, OTU!="Unknown")
print(x)

```


Parametric test

all the layers:

```{r}
library(multcomp)
library(car)
library(dplyr)

# 1. Create summary table
y_df <- psmelt(ps_x)
x <- y_df %>%
  group_by(OTU, sample_type, depth) %>% 
  summarise(
    mean = round(mean(Abundance, na.rm = TRUE) * 100, 1), 
    se = round((sd(Abundance, na.rm = TRUE) / sqrt(n())) * 100, 1),
    .groups = 'drop'
  )

# Remove Unknown OTU
x <- subset(x, OTU != "Unknown")

# Combine mean and se
x <- x %>%
  mutate(mean_se = paste0(mean, " ± ", se)) %>%
  dplyr::select(-mean, -se)

# 2. Get data for tests
x_df <- psmelt(ps_x)
x_df$OTU <- as.factor(x_df$OTU)
x_df <- subset(x_df, OTU != "Unknown")

# 3. FIRST: Compare sample_type within each depth (lowercase letters)
cld_lower <- data.frame()

for (j in unique(x_df$OTU)) {
  for (d in unique(x_df$depth)) {
    df <- x_df[x_df$OTU == j & x_df$depth == d, ]
    
    # Need at least 2 sample types
    if (length(unique(df$sample_type)) < 2) next
    
    tryCatch({
      res.aov <- aov(Abundance ~ sample_type, data = df)
      tukey <- glht(res.aov, linfct = mcp(sample_type = "Tukey"))
      cld_letters <- cld(tukey, level = 0.05)
      
      cld_df <- data.frame(
        OTU = j,
        depth = d,
        sample_type = names(cld_letters$mcletters$Letters),
        cld_lower = gsub(" ", "", cld_letters$mcletters$Letters),
        stringsAsFactors = FALSE
      )
      
      cld_lower <- rbind(cld_lower, cld_df)
    }, error = function(e) {
      # Skip if error
      message(paste("Error in lowercase for OTU", j, "depth", d, ":", e$message))
    })
  }
}

print(paste("Lowercase comparisons:", nrow(cld_lower), "rows"))

# 4. SECOND: Compare depth within each sample_type (uppercase letters)
cld_upper <- data.frame()

for (j in unique(x_df$OTU)) {
  for (st in unique(x_df$sample_type)) {
    df <- x_df[x_df$OTU == j & x_df$sample_type == st, ]
    
    # Need at least 2 depths
    if (length(unique(df$depth)) < 2) next
    
    tryCatch({
      # Convert depth to factor
      df$depth_factor <- factor(df$depth, 
                               levels = c("0...10", "10...20", "20...30", "30...40", "40..."))
      
      res.aov <- aov(Abundance ~ depth_factor, data = df)
      tukey <- glht(res.aov, linfct = mcp(depth_factor = "Tukey"))
      cld_letters <- cld(tukey, level = 0.05)
      
      cld_df <- data.frame(
        OTU = j,
        sample_type = st,
        depth = names(cld_letters$mcletters$Letters),
        cld_upper = gsub(" ", "", toupper(cld_letters$mcletters$Letters)),
        stringsAsFactors = FALSE
      )
      
      cld_upper <- rbind(cld_upper, cld_df)
    }, error = function(e) {
      # Skip if error
      message(paste("Error in uppercase for OTU", j, "sample_type", st, ":", e$message))
    })
  }
}

print(paste("Uppercase comparisons:", nrow(cld_upper), "rows"))

# 5. Check if we got results
print("Preview of cld_lower:")
print(head(cld_lower))
print("\nPreview of cld_upper:")
print(head(cld_upper))

# 6. Merge results
if (nrow(cld_lower) > 0) {
  x <- merge(x, cld_lower, by = c("OTU", "depth", "sample_type"), all.x = TRUE)
} else {
  x$cld_lower <- NA
}

if (nrow(cld_upper) > 0) {
  x <- merge(x, cld_upper, by = c("OTU", "sample_type", "depth"), all.x = TRUE)
} else {
  x$cld_upper <- NA
}

# Replace NA with empty string
x$cld_lower[is.na(x$cld_lower)] <- ""
x$cld_upper[is.na(x$cld_upper)] <- ""

# 7. Append letters directly without parentheses
x <- x %>%
  mutate(
    mean_se = case_when(
      cld_lower != "" & cld_upper != "" ~ paste(mean_se, cld_lower, cld_upper),
      cld_lower != "" ~ paste(mean_se, cld_lower),
      cld_upper != "" ~ paste(mean_se, cld_upper),
      TRUE ~ mean_se
    )
  )

# Remove the separate cld columns
x$cld_lower <- NULL
x$cld_upper <- NULL

# Arrange for better viewing
x <- x %>% arrange(OTU, depth, sample_type)

```



# 10. CAZyme specific ubstrate



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'Substrate', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Peptidoglycan <- ggplot(df, aes(x = Layer, y = Peptidoglycan, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Peptidoglycan-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Peptidoglycan

```


```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Chitin <- ggplot(df, aes(x = Layer, y = Chitin, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Chitin-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Chitin

```



```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Glucans <- ggplot(df, aes(x = Layer, y = Glucans, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Glucans-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Glucans

```



```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Cellulose <- ggplot(df, aes(x = Layer, y = Cellulose, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Cellulose-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Cellulose

```



```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Hemicellulose <- ggplot(df, aes(x = Layer, y = Hemicellulose, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Hemicellulose-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Hemicellulose

```




```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Lignin <- ggplot(df, aes(x = Layer, y = Lignin, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Lignin-derived C\ndecomposing CAZymes") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Lignin

```




# 11. Carbon target decomposing: fungi, bacteria, plant



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'C_target', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

bacteria <- ggplot(df, aes(x = Layer, y = bacteria, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "bacteria") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

bacteria

```


```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

fungi <- ggplot(df, aes(x = Layer, y = fungi, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "fungi") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

fungi

```




```{r}
names(gene_data)
```







```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

plant <- ggplot(df, aes(x = Layer, y = plant, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "plant") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

plant
```




## 11.1. Test


```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'C_target', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```




```{r}
ps_y <- subset_samples(ps_x, depth=="30...40")
ps_y
```


```{r}
library(car)

x_df <- psmelt(ps_y)

taxa <- unique(x_df$OTU)

for (i in taxa) {
 df <- filter(x_df, OTU == i)
 print(i)
 # Using leveneTest()
result = leveneTest(Abundance ~ sample_type, df)
# print the result
print(result)

}

x <- x_df %>%
  filter(OTU %in% taxa) %>% group_by(OTU, sample_type) %>% summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
print(x)

```

all are homogenous

```{r}
y_df <- psmelt(ps_x)
x <- y_df %>%
  group_by(OTU, sample_type, depth) %>% summarise(mean = round(mean(Abundance, na.rm = TRUE)*100,1), se = round((sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))) * 100,1))
x <- subset(x, OTU!="Unknown")
print(x)

```


Parametric test




```{r}
library(multcomp)
library(car)
library(dplyr)

# 1. Create summary table
y_df <- psmelt(ps_x)
x <- y_df %>%
  group_by(OTU, sample_type, depth) %>% 
  summarise(
    mean = round(mean(Abundance, na.rm = TRUE) * 100, 1), 
    se = round((sd(Abundance, na.rm = TRUE) / sqrt(n())) * 100, 1),
    .groups = 'drop'
  )

# Remove Unknown OTU
x <- subset(x, OTU != "Unknown")

# Combine mean and se
x <- x %>%
  mutate(mean_se = paste0(mean, " ± ", se)) %>%
  dplyr::select(-mean, -se)

# 2. Get data for tests
x_df <- psmelt(ps_x)
x_df$OTU <- as.factor(x_df$OTU)
x_df <- subset(x_df, OTU != "Unknown")

# 3. FIRST: Compare sample_type within each depth (lowercase letters)
cld_lower <- data.frame()

for (j in unique(x_df$OTU)) {
  for (d in unique(x_df$depth)) {
    df <- x_df[x_df$OTU == j & x_df$depth == d, ]
    
    # Need at least 2 sample types
    if (length(unique(df$sample_type)) < 2) next
    
    tryCatch({
      res.aov <- aov(Abundance ~ sample_type, data = df)
      tukey <- glht(res.aov, linfct = mcp(sample_type = "Tukey"))
      cld_letters <- cld(tukey, level = 0.05)
      
      cld_df <- data.frame(
        OTU = j,
        depth = d,
        sample_type = names(cld_letters$mcletters$Letters),
        cld_lower = gsub(" ", "", cld_letters$mcletters$Letters),
        stringsAsFactors = FALSE
      )
      
      cld_lower <- rbind(cld_lower, cld_df)
    }, error = function(e) {
      # Skip if error
      message(paste("Error in lowercase for OTU", j, "depth", d, ":", e$message))
    })
  }
}

print(paste("Lowercase comparisons:", nrow(cld_lower), "rows"))

# 4. SECOND: Compare depth within each sample_type (uppercase letters)
cld_upper <- data.frame()

for (j in unique(x_df$OTU)) {
  for (st in unique(x_df$sample_type)) {
    df <- x_df[x_df$OTU == j & x_df$sample_type == st, ]
    
    # Need at least 2 depths
    if (length(unique(df$depth)) < 2) next
    
    tryCatch({
      # Convert depth to factor
      df$depth_factor <- factor(df$depth, 
                               levels = c("0...10", "10...20", "20...30", "30...40", "40..."))
      
      res.aov <- aov(Abundance ~ depth_factor, data = df)
      tukey <- glht(res.aov, linfct = mcp(depth_factor = "Tukey"))
      cld_letters <- cld(tukey, level = 0.05)
      
      cld_df <- data.frame(
        OTU = j,
        sample_type = st,
        depth = names(cld_letters$mcletters$Letters),
        cld_upper = gsub(" ", "", toupper(cld_letters$mcletters$Letters)),
        stringsAsFactors = FALSE
      )
      
      cld_upper <- rbind(cld_upper, cld_df)
    }, error = function(e) {
      # Skip if error
      message(paste("Error in uppercase for OTU", j, "sample_type", st, ":", e$message))
    })
  }
}

print(paste("Uppercase comparisons:", nrow(cld_upper), "rows"))

# 5. Check if we got results
print("Preview of cld_lower:")
print(head(cld_lower))
print("\nPreview of cld_upper:")
print(head(cld_upper))

# 6. Merge results
if (nrow(cld_lower) > 0) {
  x <- merge(x, cld_lower, by = c("OTU", "depth", "sample_type"), all.x = TRUE)
} else {
  x$cld_lower <- NA
}

if (nrow(cld_upper) > 0) {
  x <- merge(x, cld_upper, by = c("OTU", "sample_type", "depth"), all.x = TRUE)
} else {
  x$cld_upper <- NA
}

# Replace NA with empty string
x$cld_lower[is.na(x$cld_lower)] <- ""
x$cld_upper[is.na(x$cld_upper)] <- ""

# 7. Append letters directly without parentheses
x <- x %>%
  mutate(
    mean_se = case_when(
      cld_lower != "" & cld_upper != "" ~ paste(mean_se, cld_lower, cld_upper),
      cld_lower != "" ~ paste(mean_se, cld_lower),
      cld_upper != "" ~ paste(mean_se, cld_upper),
      TRUE ~ mean_se
    )
  )

# Remove the separate cld columns
x$cld_lower <- NULL
x$cld_upper <- NULL

# Arrange for better viewing
x <- x %>% arrange(OTU, depth, sample_type)

```




# 12. LINEAR MODEL


```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'Derived', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```


```{r}
y_df <- psmelt(ps_x)
y_df <- y_df[, (c("OTU", "Abundance", "sampleID"))]
df <- y_df %>% 
  pivot_wider(
    names_from = OTU, 
    values_from = Abundance,
    values_fill = 0  # Optional: fill missing values with 0
  )
meta <- meta(ps_x)
meta <- left_join(meta, df, by="sampleID")
names(meta)
```

Complete analysis WITH TYPE ii ANOVA


without forest

```{r}
library(lmtest)
library(sandwich)
library(car)
library(effectsize)
library(performance)

meta_nf <- subset(meta, sample_type!="forest")
# 1. Main analysis with effect sizes
m1 <- lm(C_g_per_kg ~ depth + Plant, meta_nf)
summary(m1)
aov_t2 <- Anova(m1, type = "II")
effects <- eta_squared(aov_t2, partial = TRUE)
print(effects)

# Key diagnostics to report:
check_normality(m1)
check_heteroscedasticity(m1)
check_outliers(m1)

# 2. Robustness check
robust_anova <- Anova(m1, type = "II", vcov = vcovHC(m1, type = "HC3"))
robust_coef <- coeftest(m1, vcov = vcovHC(m1, type = "HC3"))

# 3. Compare
print(robust_anova)
print(robust_coef)
```



# 13. Alternative figures

Plant and Microbial derived C decomposing



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'Derived', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```


```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


Reshape df

```{r}
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

df$plant_microbe_ratio <- df$Plant / df$Microbial
df$plant_minus_microbe <- df$Plant - df$Microbial

df_long <- df %>%
  pivot_longer(
    cols = c(Microbial, Plant),  # Columns to combine
    names_to = "Derived",  # New column for GH/GT
    values_to = "Proportion"  # New column for values
  )
```






```{r}
fig <- df_long %>%
  group_by(sample_type, depth_numerical, Derived) %>%
  summarise(
    mean = mean(Proportion, na.rm = TRUE),
    se = sd(Proportion, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  facet_wrap(~ Derived, 
             scales = "free_x",
             labeller = labeller(Derived = c(
               "Microbial" = "Microbial-derived C\ndecomposing CAZymes",
               "Plant" = "Plant-derived C\ndecomposing CAZymes"
             ))) +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "Proportion",
    color = "Treatment"
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 16),
    legend.title=element_text(size = 16),
    plot.title = element_text(size = 16, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    strip.text = element_text(size = 16)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  )

print(fig)
```






C target decomposing



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'C_target', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```


```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


Reshape df

```{r}
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

df_long <- df %>%
  pivot_longer(
    cols = c(bacteria, fungi, plant),  # Columns to combine
    names_to = "Derived",  # New column for GH/GT
    values_to = "Proportion"  # New column for values
  )
```






```{r}
fig <- df_long %>%
  group_by(sample_type, depth_numerical, Derived) %>%
  summarise(
    mean = mean(Proportion, na.rm = TRUE),
    se = sd(Proportion, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  facet_wrap(~ Derived, 
             scales = "free_x",
             labeller = labeller(Derived = c(
               "bacteria" = "Bacterial-derived C",
               "fungi" = "Fungi-derived C", 
               "plant" = "Plant-derived C"
             ))) +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "Proportion",
    color = "Treatment"
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    strip.text = element_text(size = 12)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  )

print(fig)
```









Plant and Microbial derived C decomposing



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'Substrate', detection = 0/100, prevalence = 0/140)
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x
```



```{r}
ps_y <- ps_x
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```


```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


Reshape df

```{r}
df <- merge(metadata, gene_data, by=0, all.x=TRUE)


df_long <- df %>%
  pivot_longer(
    cols = c(Peptidoglycan, Chitin, Glucans, Cellulose, Hemicellulose, Lignin),  # Columns to combine
    names_to = "Substrate",  # New column for GH/GT
    values_to = "Proportion"  # New column for values
  )
```









```{r}
df_long$Substrate <- factor(df_long$Substrate, levels = c("Lignin", "Hemicellulose", "Cellulose", "Peptidoglycan", "Chitin", "Glucans"))

fig <- df_long %>%
  group_by(sample_type, depth_numerical, Substrate) %>%
  summarise(
    mean = mean(Proportion, na.rm = TRUE),
    se = sd(Proportion, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  facet_wrap(~ Substrate, 
             scales = "free_x",
             labeller = labeller(Substrate = c(
               "Peptidoglycan" = "Peptidoglycan (bacterial)",
               "Chitin" = "Chitin (fungal)",
               "Glucans" = "Glucans (fungal)"
             ))) +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "Proportion of CAZyme genes targeting different C fractions",
    color = "Treatment"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),    # This now overrides cowplot
    legend.justification = "center",  # This centers the legend
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 20, face = "bold"),
    legend.text = element_text(size = 20),
    plot.title = element_text(size = 20, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    strip.text = element_text(size = 20)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  )

print(fig)
```





CAZyme family figures

```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- aggregate_rare(ps_RA, level = 'family', detection = 0/100, prevalence = 0/140)# change the level accordingly: "Derived"   "C_target"  "Substrate" "family"    "enzyme"
ps_x


tax <- as.data.frame(tax_table(ps_x))

meta <- meta(ps)

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_x) <- sample_data(meta)

ps_x

df <- psmelt(ps_x)
```


GT


```{r}
GT <- df %>%
  filter(OTU %in% c("GT")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "GT",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(GT)
```




GH


```{r}
GH <- df %>%
  filter(OTU %in% c("GH")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "GH",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(GH)
```



 AA


```{r}
AA <- df %>%
  filter(OTU %in% c("AA")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, shape = 15, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "AA",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(AA)
```



CE


```{r}
CE <- df %>%
  filter(OTU %in% c("CE")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, shape = 15, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "CE",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(CE)
```




 PL


```{r}
PL <- df %>%
  filter(OTU %in% c("PL")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, shape = 15, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "PL",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(PL)
```


CBM


```{r}
CBM <- df %>%
  filter(OTU %in% c("CBM")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "CBM",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(CBM)
```




SLH


```{r}
SLH <- df %>%
  filter(OTU %in% c("SLH")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "SLH",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(SLH)
```




Cohesin


```{r}
Cohesin <- df %>%
  filter(OTU %in% c("cohesin")) %>%
  group_by(sample_type, depth_numerical, OTU) %>%
  summarise(
    mean = mean(Abundance, na.rm = TRUE),
    se = sd(Abundance, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "Cohesin",
    color = NULL  # Remove legend title
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles only
    legend.text = element_text(size = 14),  # No bold for legend text
    legend.title = element_blank(),  # Remove legend title (alternative method)
    plot.title = element_text(size = 18, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5)
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) +
  guides(color = guide_legend(override.aes = list(shape = 15)))

print(Cohesin)
```




Combine important CAZyme family figures



```{r}

library(ggpubr)

figure1 <-  ggarrange(GH, GT + rremove("ylab") + rremove("y.text") + rremove("y.ticks"), widths = c(1, 0.80, 0.80), hjust = c(-0.5, 0.5, 0.5),
                    align = "h", ncol = 3, nrow = 1, common.legend = TRUE, legend = "top")

figure2 <-  ggarrange(AA, CE + rremove("ylab") + rremove("y.text") + rremove("y.ticks"), PL + rremove("ylab") + rremove("y.text") + rremove("y.ticks"),
                    widths = c(1, 0.80, 0.80), hjust = c(-0.5, 0.5, 0.5),
                    align = "h", ncol = 3, nrow = 1, legend = "none")

figure <- ggarrange(figure1, figure2,
                   heights = c(1, 1),
                    align = "h", ncol = 1, nrow = 2)

#figure <- annotate_figure(figure, left = text_grob("CAZymes decomposing C derived from",size = 14, rot = 90, face = "bold"))

figure
```




```{r}
rm(list = ls())
```





