---
title: "Untitled"
output: html_document
date: "2026-05-18"
---



# SEM

We will do a sequential structural equation model (SEM) to assess effectors of soil carbon stocks along the soil profile

I will test with different microbial indicators: KEGG, CAZyme, and taxonomic (read-based) compositions, for all of which the first axis of PCoA will be used.

# 1. Load data


```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)

setwd('C:\\Users\\lehakkin\\RStudio_computer')

# let's firs load KEGG data
load(file = 'KOfam_metabolism_ps_COVERAGE_FINAL_PNSCO2')
ps

ps_RA <- transform(ps, "compositional")

meta <- meta(ps)

```


## 1.1. remove forest

```{r}
# subset samples
ps_RA <- subset_samples(ps_RA, sample_type != "forest")


# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_RA),
              MARGIN = ifelse(taxa_are_rows(ps_RA), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_RA = prune_taxa((prev0 > 0), ps_RA)
ps_RA

```


```{r}
ps_RA_KEGG <- ps_RA
ps_RA_KEGG
```




```{r}
setwd('C:\\Users\\lehakkin\\RStudio_computer')
load(file = 'CAZyme_ps_COVERAGE_FINAL')#ps
ps

ps_CAZyme <- ps

rm(ps)
```



CAZyme SUBSTRATES

```{r}
substrates <- read.csv2(file = "C:\\Users\\lehakkin\\RStudio_computer\\CAZyme_Substrates_from_Huan_et_al_2024.csv")
tax <- as.data.frame(tax_table(ps_CAZyme))
tax <- left_join(tax, substrates, by=c("enzyme"="CAZyme_family"))
tax <- as.data.frame(tax)
tax <- tax[, c(3:5, 1,2)]
row.names(tax) <- tax$enzyme

tax_table(ps_CAZyme) <- tax_table(as.matrix(tax))
ps_CAZyme
rank_names(ps_CAZyme)
```


```{r}

# subset samples
ps_CAZyme <- subset_samples(ps_CAZyme, sample_type != "forest")


# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_CAZyme),
              MARGIN = ifelse(taxa_are_rows(ps_CAZyme), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_CAZyme = prune_taxa((prev0 > 0), ps_CAZyme)
ps_CAZyme

ps_CAZyme_RA <- transform(ps_CAZyme, "compositional")

```

 either

```{r}
# Step 1: Aggregate to Substrate level
ps_y <- aggregate_rare(ps_CAZyme_RA, level = "Substrate", detection = 0, prevalence = 0)

```


or!!!!!!!!!!

```{r}
#ps_y <- aggregate_rare(ps_CAZyme_RA, level = 'Derived', detection = 0/100, prevalence = 0/140)
#ps_y
```





```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```







```{r}
#left join
meta <- merge(metadata, gene_data, by=0, all.x=TRUE)
```




```{r}

colnames(meta)[which(names(meta) == "pH_H2O")] <- "pH"
colnames(meta)[which(names(meta) == "EC_uScm")] <- "EC"
colnames(meta)[which(names(meta) == "C_g_per_kg")] <- "C"
colnames(meta)[which(names(meta) == "N_gkg")] <- "N"
colnames(meta)[which(names(meta) == "TP_gkg")] <- "P"
colnames(meta)[which(names(meta) == "Alox_mmolkg")] <- "Al-ox"
colnames(meta)[which(names(meta) == "Feox_mmolkg")] <- "Fe-ox"
colnames(meta)[which(names(meta) == "oxides_mmolkg")] <- "Oxides"
colnames(meta)[which(names(meta) == "PH2O_mgkg")] <- "P-H2O"
colnames(meta)[which(names(meta) == "Porg_mgkg")] <- "P-org"
colnames(meta)[which(names(meta) == "DOC_mgkg")] <- "DOC"
colnames(meta)[which(names(meta) == "Pinorg_mgkg")] <- "P-inorg"
colnames(meta)[which(names(meta) == "C_per_N")] <- "C/N"
colnames(meta)[which(names(meta) == "log_root")] <- "Root biomass"
colnames(meta)[which(names(meta) == "depth_numerical")] <- "Depth (numerical)"



variables <- c("pH", "EC", "C", "N", "P", "Al-ox", "Fe-ox", "Oxides", "P-H2O", "P-org", "DOC", "P-inorg", "C/N", "Root biomass", "Depth (numerical)")

```



```{r}
MyPalette2 <- c(forest = "#1167b1", Meadow = "#fbc02d", Organic = "#8a8a8a", Conventional =  "#b71c1c")

meta$soil_layer <- meta$depth
meta$soil_layer[meta$soil_layer=="0...10"]<- "0-10 cm"
meta$soil_layer[meta$soil_layer=="10...20"]<- "10-20 cm"
meta$soil_layer[meta$soil_layer=="20...30"]<- "20-30 cm"
meta$soil_layer[meta$soil_layer=="30...40"]<- "30-40 cm"
meta$soil_layer[meta$soil_layer=="40..."]<- "40-80 cm"


#rename column by name
colnames(meta)[colnames(meta) == 'sample_type'] <- 'treatment'

# Convert factor to character
meta$treatment <- as.character(meta$treatment)

# Now do your replacements
meta$treatment[meta$treatment == "meadow"] <- "Meadow"
meta$treatment[meta$treatment == "organic"] <- "Organic"
meta$treatment[meta$treatment == "conventional"] <- "Conventional"

# Optional: Convert back to factor if needed
meta$treatment <- factor(meta$treatment)

meta$treatment <- factor(meta$treatment, levels = c("Meadow", "Organic", "Conventional"))

```




# 2. Extract PCoA axes

```{r}
# Save Bray-Curtis dissimilarity matrices as separate objects

CAZyme_dist <- distance(ps_CAZyme_RA, method = "bray")
KEGG_dist <- distance(ps_RA_KEGG, method = "bray")



# Function to extract PCoA axes SEPARATELY
extract_pcoa <- function(ps_object, prefix, n_axes = 3) {
  dist_matrix <- distance(ps_object, method = "bray")
  pcoa_result <- cmdscale(dist_matrix, k = n_axes, eig = TRUE)
  axes <- as.data.frame(pcoa_result$points)
  colnames(axes) <- paste0(prefix, "_PCoA", 1:n_axes)
  return(axes)
}

# Extract PCoA axes for all communities SEPARATELY (using the same function)
CAZyme_pcoa <- extract_pcoa(ps_CAZyme_RA, "CAZyme", n_axes = 3)
KEGG_pcoa <- extract_pcoa(ps_RA_KEGG, "KEGG", n_axes = 3)


# Prepare combined dataset
combined_data <- meta %>%
  # Bind microbiome data
  bind_cols(CAZyme_pcoa, KEGG_pcoa) %>%
  mutate(
    # 1. TREATMENT as factor with Meadow as reference
    treatment = factor(treatment, 
                      levels = c("Meadow", "Organic", "Conventional")),
    
    # 2. Depth as numeric
    depth_numeric = case_when(
      depth == "0...10" ~ 5,
      depth == "10...20" ~ 15,
      depth == "20...30" ~ 25,
      depth == "30...40" ~ 35,
      depth == "40..." ~ 60,
      TRUE ~ as.numeric(gsub("[^0-9]", "", depth))
    ),
    
    # 3. Standardize soil properties and PCoA axes
    across(c(C, pH, N, P, `Al-ox`, 
             `Fe-ox`, `Root biomass`, `C/N`, EC, `P-org`, DOC),
           ~ scale(.x)[,1], .names = "{.col}_z"),
    
    # 4. Standardize PCoA axes SEPARATELY
    across(starts_with("CAZyme_"), ~ scale(.x)[,1], .names = "{.col}_z"),
    across(starts_with("KEGG_"), ~ scale(.x)[,1], .names = "{.col}_z")
  )

# Check column names to confirm separation
grep("CAZyme|KEGG", names(combined_data), value = TRUE)

```





# 3. SEM

## 3.1.  piecewise SEM 

Preliminary analysis revealed that CAZymes composition correlated with soil carbon stocks the most and is thus kept in the final SEM

```{r}

library(lme4)
library(piecewiseSEM)

m.sem <- psem(
  lmer(pH ~ Organic + Conventional + depth_numerical + (1|plot), data = df),
  lmer(Al_ox ~ pH + Organic + Conventional + depth_numerical + (1|plot), data = df),
  lmer(Fe_ox ~ pH + Organic + Conventional + (1|plot), data = df),
  lmer(log_root ~ Organic + Conventional + depth_numerical + (1|plot), data = df),
  lm(CAZyme_PCoA1 ~ depth_numerical + pH + Fe_ox, data = df),
  lm(C ~ CAZyme_PCoA1 + depth_numerical + Al_ox + log_root, data = df),
  data = df
)

summary(m.sem)

```

Diagnostics

```{r}
# Access each model by index
model1 <- m.sem[[1]]  # pH
model2 <- m.sem[[2]]  # Al_ox
model3 <- m.sem[[3]]  # Fe_ox
model4 <- m.sem[[4]]  # log_root
model5 <- m.sem[[5]]  # CAZyme_PCoA1
model6 <- m.sem[[6]]  # C

# Or create a list
models <- list(
  pH = m.sem[[1]],
  Al_ox = m.sem[[2]],
  Fe_ox = m.sem[[3]],
  log_root = m.sem[[4]],
  CAZyme_PCoA1 = m.sem[[5]],
  C = m.sem[[6]]
)

# Check class of each
sapply(models, class)
```


```{r}
library(lmtest)
library(performance)
library(car)

# Define model names
model_names <- c("pH", "Al_ox", "Fe_ox", "log_root", "CAZyme_PCoA1", "C")

for(i in 1:6) {
  model <- m.sem[[i]]
  model_name <- model_names[i]
  
  cat("\n", paste(rep("=", 70), collapse = ""), "\n")
  cat("MODEL:", model_name, "\n")
  cat(paste(rep("=", 70), collapse = ""), "\n")
  
  # Summary
  print(summary(model))
  
  # Residuals
  resid <- residuals(model)
  
  # Normality (Shapiro-Wilk)
  shapiro_p <- shapiro.test(resid)$p.value
  cat("\nShapiro-Wilk p =", round(shapiro_p, 4),
      ifelse(shapiro_p > 0.05, " ✓ normal", " ⚠ non-normal"))
  
  # Autocorrelation (Durbin-Watson)
  if(length(resid) > 3) {
    dw_p <- dwtest(resid ~ 1)$p.value
    cat("\nDurbin-Watson p =", round(dw_p, 4),
        ifelse(dw_p > 0.05, " ✓ no autocorrelation", " ⚠ autocorrelation"))
  }
  
  # R²
  if(inherits(model, "lmerMod")) {
    r2 <- r2_nakagawa(model)
    cat("\nMarginal R² =", round(r2$R2_marginal, 4))
    cat("\nConditional R² =", round(r2$R2_conditional, 4))
    
    # Random effect variance
    vc <- VarCorr(model)
    plot_var <- as.numeric(vc$plot[1,1])
    cat("\nPlot variance =", round(plot_var, 6))
  } else {
    cat("\nR² =", round(summary(model)$r.squared, 4))
  }
  
  # Influential points (for lm only)
  if(inherits(model, "lm") & !inherits(model, "lmerMod")) {
    cooks <- cooks.distance(model)
    threshold <- 4 / length(cooks)
    n_inf <- sum(cooks > threshold)
    cat("\nInfluential points (Cook's D >", round(threshold, 3), "):", n_inf)
    if(n_inf > 0) {
      cat(" -", paste(which(cooks > threshold), collapse = ", "))
    }
  }
  
  cat("\n")
}
```



#### 3.4.1. Extract standardized coefficients

```{r}
library(dplyr)

std_coefs <- coefs(m.sem, standardize="scale")

std_coefs <- std_coefs[,1:8]

names(std_coefs) <- c(
"Response",
"Predictor",
"Estimate",
"Std.Error",
"DF",
"Crit.Value",
"P.Value",
"Std.Estimate"
)

get_coef <- function(resp, pred){

  val <- std_coefs %>%
    filter(Response==resp,
           Predictor==pred) %>%
    pull(Std.Estimate)

  if(length(val)==0) return(0)

  val
}
```


#### 3.4.2. Define ALL valid indirect paths explicitly


```{r}
ind_org <- c(

# root
get_coef("log_root","Organic") *
get_coef("C","log_root"),

# Al paths
get_coef("pH","Organic") *
get_coef("Al_ox","pH") *
get_coef("C","Al_ox"),

get_coef("Al_ox","Organic") *
get_coef("C","Al_ox"),

# CAZyme paths
get_coef("pH","Organic") *
get_coef("CAZyme_PCoA1","pH") *
get_coef("C","CAZyme_PCoA1"),

get_coef("Fe_ox","Organic") *
get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1"),

get_coef("pH","Organic") *
get_coef("Fe_ox","pH") *
get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1")

)

indirect_Organic <- sum(ind_org)
```



```{r}
ind_conv <- c(

get_coef("log_root","Conventional") *
get_coef("C","log_root"),

get_coef("pH","Conventional") *
get_coef("Al_ox","pH") *
get_coef("C","Al_ox"),

get_coef("Al_ox","Conventional") *
get_coef("C","Al_ox"),

get_coef("pH","Conventional") *
get_coef("CAZyme_PCoA1","pH") *
get_coef("C","CAZyme_PCoA1"),

get_coef("Fe_ox","Conventional") *
get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1"),

get_coef("pH","Conventional") *
get_coef("Fe_ox","pH") *
get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1")

)

indirect_Conventional <- sum(ind_conv)
```


```{r}
ind_depth <- c(

get_coef("log_root","depth_numerical") *
get_coef("C","log_root"),

get_coef("pH","depth_numerical") *
get_coef("Al_ox","pH") *
get_coef("C","Al_ox"),

get_coef("Al_ox","depth_numerical") *
get_coef("C","Al_ox"),

get_coef("depth_numerical","CAZyme_PCoA1") *
get_coef("C","CAZyme_PCoA1"),

get_coef("pH","depth_numerical") *
get_coef("CAZyme_PCoA1","pH") *
get_coef("C","CAZyme_PCoA1"),

get_coef("pH","depth_numerical") *
get_coef("Fe_ox","pH") *
get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1")

)

indirect_depth <- sum(ind_depth)
```


```{r}
ind_pH <- c(

get_coef("Al_ox","pH") *
get_coef("C","Al_ox"),

get_coef("CAZyme_PCoA1","pH") *
get_coef("C","CAZyme_PCoA1"),

get_coef("Fe_ox","pH") *
get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1")

)

indirect_pH <- sum(ind_pH)
```


```{r}
indirect_Fe <-

get_coef("CAZyme_PCoA1","Fe_ox") *
get_coef("C","CAZyme_PCoA1")
```


#### 3.4.3. Direct effects

```{r}
direct_effects <- std_coefs %>%
  dplyr::filter(Response=="C") %>%
  dplyr::select(Predictor,
         Direct = Std.Estimate)
```



#### 3.4.4. Combine indirect + direct

```{r}
indirect_df <- data.frame(

Predictor = c(
"Organic",
"Conventional",
"depth_numerical",
"pH",
"Fe_ox"
),

Indirect = c(
indirect_Organic,
indirect_Conventional,
indirect_depth,
indirect_pH,
indirect_Fe
)

)
```


#### 3.4.5. Total effects

```{r}
effects_all <- full_join(
direct_effects,
indirect_df,
by="Predictor"
)

effects_all[is.na(effects_all)] <- 0

effects_all <- effects_all %>%

mutate(
Total = Direct + Indirect
)
```



#### 3.4.6. Plot total effects




```{r}
library(ggplot2)

effects_all %>%

mutate(
color = ifelse(Total>0,"Positive","Negative")
) %>%

ggplot(aes(
x=reorder(Predictor,Total),
y=Total,
fill=color
))+

geom_bar(stat="identity")+

coord_flip()+

scale_fill_manual(
values=c(
"Positive"="#0070C0",
"Negative"="#C00000"
)
)+

geom_hline(
yintercept=0,
linetype="dashed"
)+

labs(
y="Total standardized effects",
x=NULL
)+

theme_minimal()+

guides(fill="none")+

# THIS fixes the labels
scale_x_discrete(labels = function(x) {
  x[x == "Fe_ox"] <- "Fe-ox"
  x[x == "Al_ox"] <- "Al-ox"
  x[x == "depth_numerical"] <- "Soil\ndepth"
  x[x == "log_root"] <- "Root\nbiomass"
  x[x == "CAZyme_PCoA1"] <- "CAZyme\nPCoA1"
  x[x == "Organic"] <- "Organic\ntreatment"
  x[x == "Conventional"] <- "Conventional\ntreatment"
  x
})
```


#### 3.4.7. Plot direct effects

```{r}
library(ggplot2)
library(dplyr)

custom_labels <- c(

"Organic" = "Organic\ntreatment",
"Conventional" = "Conventional\ntreatment",
"depth_numerical" = "Soil\ndepth",

"pH" = "pH",

"Fe_ox" = "Fe-ox",
"Al_ox" = "Al-ox",

"CAZyme_PCoA1" = "CAZyme\nPCoA1",

"log_root" = "Root\nbiomass"

)

plot_direct <- direct_effects %>%

mutate(

Predictor_label = custom_labels[Predictor],

color = ifelse(
Direct >= 0,
"Positive",
"Negative"
)

) %>%

ggplot(

aes(
x = reorder(Predictor_label, Direct),
y = Direct,
fill = color
)

) +

geom_bar(stat = "identity") +

coord_flip() +

scale_fill_manual(
values = c(
"Positive" = "#0070C0",
"Negative" = "#C00000"
)
) +

geom_hline(
yintercept = 0,
linetype = "dashed"
) +

labs(
y = "Direct standardized effects",
x = NULL
) +

theme_minimal() +

theme(

plot.background = element_rect(fill = "transparent", color = NA),
panel.background = element_rect(fill = "transparent", color = NA),
legend.background = element_rect(fill = "transparent", color = NA),

axis.text.x = element_text(size = 16),
axis.text.y = element_text(size = 16),

axis.title.x = element_text(size = 20),
axis.title.y = element_text(size = 20)

) +

guides(fill = "none") +

scale_y_continuous(
limits = c(-0.3, 0.4),
breaks = seq(-0.3, 0.4, by = 0.1),
labels = function(x) sprintf("%.1f", x)
)

plot_direct

```


#### 3.4.8. lot indirect effects

```{r}
plot_indirect <- indirect_df %>%

mutate(
color = ifelse(Indirect>0,"Positive","Negative")
) %>%

ggplot(aes(
x=reorder(Predictor,Indirect),
y=Indirect,
fill=color
))+

geom_bar(stat="identity")+

coord_flip()+

scale_fill_manual(
values=c(
"Positive"="#0070C0",
"Negative"="#C00000"
)
)+

geom_hline(
yintercept=0,
linetype="dashed"
)+

labs(
y="Indirect standardized effects",
x=NULL
)+

theme_minimal()+

guides(fill="none")+

# THIS is the only added part
scale_x_discrete(labels = c(
  "Organic" = "Organic\ntreatment",
  "Conventional" = "Conventional\ntreatment",
  "depth_numerical" = "Soil\ndepth",
  "pH" = "pH",
  "Fe_ox" = "Fe-ox",
  "Al_ox" = "Al-ox",
  "CAZyme_PCoA1" = "CAZyme\nPCoA1",
  "log_root" = "Root\nbiomass"
))

plot_indirect
```



#### 3.4.9. Stacked diret and indirect

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Convert to long format
effects_long <- effects_all %>%
  
  dplyr::select(Predictor, Direct, Indirect) %>%
  
  pivot_longer(
    cols = c(Direct, Indirect),
    names_to = "Effect_type",
    values_to = "Value"
  ) %>%
  
  mutate(
    
    Direction = ifelse(Value >= 0,
                       "Positive",
                       "Negative"),
    
    Fill_type = case_when(
      Direction == "Positive" & Effect_type == "Direct"
        ~ "Pos_Direct",
      
      Direction == "Positive" & Effect_type == "Indirect"
        ~ "Pos_Indirect",
      
      Direction == "Negative" & Effect_type == "Direct"
        ~ "Neg_Direct",
      
      Direction == "Negative" & Effect_type == "Indirect"
        ~ "Neg_Indirect"
    )
    
  )

plot_total <- ggplot(
  effects_long,
  
  aes(
    x = reorder(Predictor, Value),
    y = Value,
    fill = Fill_type
  )
) +

geom_bar(stat="identity") +

coord_flip() +

# Color system: red/blue + light/dark
scale_fill_manual(
  values = c(
    
    "Pos_Direct"   = "#0070C0",  # dark blue
    "Pos_Indirect" = "#9DC3E6",  # light blue
    
    "Neg_Direct"   = "#C00000",  # dark red
    "Neg_Indirect" = "#F4B6B6"   # light red
    
  ),
  
  labels = c(
    "Pos_Direct" = "Direct (+)",
    "Pos_Indirect" = "Indirect (+)",
    "Neg_Direct" = "Direct (−)",
    "Neg_Indirect" = "Indirect (−)"
  )
) +

geom_hline(
  yintercept=0,
  linetype="dashed"
) +

labs(
  y="Total standardized effects",
  x=NULL,
  fill=NULL
) +

theme_minimal() +

theme(
  plot.background = element_rect(fill="transparent", color=NA),
  panel.background = element_rect(fill="transparent", color=NA),
  legend.background = element_rect(fill="transparent", color=NA),
  
  legend.position = "top", 
  legend.text = element_text(size=16),
  
  axis.text.x = element_text(size=16),
  axis.text.y = element_text(size=16, color = "black"),
  
  axis.title.x = element_text(size=20),
  axis.title.y = element_text(size=20)
) +
  guides(fill = guide_legend(nrow = 2)) +
  
# Your original label fixes
scale_x_discrete(labels = function(x) {
  x[x == "Fe_ox"] <- "Fe-ox"
  x[x == "Al_ox"] <- "Al-ox"
  x[x == "depth_numerical"] <- "Soil\ndepth"
  x[x == "log_root"] <- "Root\nbiomass"
  x[x == "CAZyme_PCoA1"] <- "CAZyme\nPCoA1"
  x[x == "Organic"] <- "Organic\ntreatment"
  x[x == "Conventional"] <- "Conventional\ntreatment"
  x
})

plot_total
```



# 4. Figures


```{r}
MyPalette <- c(forest = "#1167b1", Meadow = "#fbc02d", Organic = "#8a8a8a", Conventional =  "#b71c1c")

library("cowplot")

```



## 4.1. C in treatment layers

```{r}
C_g_per_kg <- meta %>%
  group_by(treatment, `Depth (numerical)`) %>%
  summarise(mean = mean(C, na.rm = TRUE), se = (sd(C, na.rm = TRUE)/sqrt(length((C))))) %>%
  ungroup() %>%
  ggplot(aes(y=mean, `Depth (numerical)`, color=treatment)) +
  geom_line(linetype = "dashed") +
  geom_point(size=2, position=position_dodge(1.5)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5, position=position_dodge(1.5)) + theme(plot.title = element_text(size = 20, hjust = 0.5)) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=20),
        axis.title = element_text(size=20),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        title = element_text(size=20)) +
  scale_y_continuous(name="C (g/kg)") + labs(x = NULL) + theme(legend.title = element_blank()) + scale_colour_manual(values = MyPalette2) + scale_x_reverse(breaks = c(60, 35, 25, 15, 5), labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")) + coord_flip()

C_g_per_kg
```





```{r}

p <- ggplot(meta, aes(y = C, 
                      x = `Al-ox`, 
                      color = treatment,    # color by treatment
                      shape = depth)) +     # shape by depth
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment)) +
  theme_minimal() +
  labs(y = "C", x = "Al-ox (mmol/kg)")

p

```



```{r}

p <- ggplot(meta, aes(y = C, 
                      x = `Fe-ox`, 
                      color = treatment,    # color by treatment
                      shape = depth)) +     # shape by depth
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, aes(group = treatment)) +
  theme_minimal() +
  labs(y = "C (g/kg)", y = "Fe-ox (mmol/kg)")

p

```




# 5. CAZyme groups correlating with CAZyme_PCoA1


```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)

# Step 1: Aggregate to Substrate level
ps_Substrate <- aggregate_rare(ps_CAZyme_RA, level = "Substrate", detection = 0, prevalence = 0)

# Step 2: Extract Substrate-level abundances
otu_Substrate <- as.data.frame(otu_table(ps_Substrate))
if(taxa_are_rows(ps_Substrate)) {
  otu_Substrate <- as.data.frame(t(otu_Substrate))
}

# Step 3: Get Substrate names
Substrate_names <- tax_table(ps_Substrate)[, "Substrate"]

# Step 4: Get PCoA1 scores from your meta data
pcoa1_scores <- df$CAZyme_PCoA1[match(rownames(otu_Substrate), df$sampleID)]

# Step 5: Calculate correlations for each Substrate
Substrate_correlations <- data.frame(
  Substrate = Substrate_names,
  correlation = NA,
  p_value = NA
)

for(i in 1:ncol(otu_Substrate)) {
  if(sd(otu_Substrate[, i]) > 0) {
    test <- cor.test(pcoa1_scores, otu_Substrate[, i], method = "spearman")
    Substrate_correlations$correlation[i] <- test$estimate
    Substrate_correlations$p_value[i] <- test$p.value
  }
}

# Step 6: Clean up and adjust p-values
Substrate_correlations <- Substrate_correlations[!is.na(Substrate_correlations$correlation), ]
Substrate_correlations$p_adjusted <- p.adjust(Substrate_correlations$p_value, method = "fdr")
Substrate_correlations <- Substrate_correlations[order(-abs(Substrate_correlations$correlation)), ]

# Step 7: Add significance stars
Substrate_correlations <- Substrate_correlations %>%
  mutate(
    significance = case_when(
      p_adjusted < 0.001 ~ "***",
      p_adjusted < 0.01 ~ "**",
      p_adjusted < 0.05 ~ "*",
      p_adjusted < 0.1 ~ ".",
      TRUE ~ ""
    ),
    Substrate = factor(Substrate, levels = Substrate[order(correlation)]),
    direction = ifelse(correlation > 0, "Positive", "Negative")
  )

# View results
print(Substrate_correlations)

# Remove Unknown if present
Substrate_correlations <- subset(Substrate_correlations, Substrate != "Unknown" & !is.na(Substrate))

# Create the plot
p4 <- ggplot(Substrate_correlations, aes(x = correlation, y = Substrate)) +
  geom_col(width = 0.7, fill = "gray50") +  # Set a specific color for bars
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30", size = 0.8) +
  theme_minimal() +
  labs(
    x = "Spearman correlation with CAZyme PCoA1",
    y = "CAZymes decomposing C derived from:",
    title = NULL,
    fill = "Direction"
  ) +
  theme(
    axis.text.y = element_text(size = 14, face = "bold", color = "black"),  # Match bars
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14, color = "black"),  # Match bars
    plot.title = element_blank(),
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(hjust = 0, face = "italic", size = 10),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  ) +
  # Add significance stars
  geom_text(aes(label = significance, 
                x = ifelse(correlation > 0, correlation + 0.03, correlation - 0.03)),
            hjust = ifelse(Substrate_correlations$correlation > 0, 0, 1),
            size = 6) +
  scale_x_continuous(
    limits = c(-1, 1),
    breaks = seq(-1, 1, by = 0.2),
    expand = c(0.02, 0.02)
  ) + # Manually set the axis labels
  scale_y_discrete(labels = c(
    "Chitin" = "Chitin\n(fungal)",
    "Peptidoglycan" = "Peptidoglycan\n(bacterial)",
    "Cellulose" = "Cellulose\n(plant)",
    "Hemicellulose" = "Hemicellulose\n(plant)",
    "Glucans" = "Glucans\n(fungal)",
    "Lignin" = "Lignin\n(plant")) 

print(p4)


```





# 6. Correlations with C

## 6.1. Root biomass

```{r}
library(ggpmisc)

meta %>% 
    ggplot(aes(x = `Root biomass`, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Adding the equation and R-squared value
  stat_poly_eq(formula = y ~ x,
               aes(label = paste(after_stat(eq.label), 
                                 after_stat(rr.label), 
                                 sep = "~~~")),
               parse = TRUE) +
   labs(x = "Root biomass (log10)",
       y =  "C (g/kg)") +
  theme_test()

```


## 6.2. Depth

```{r}
library(ggpmisc)

meta %>% 
   ggplot(aes(x = `Depth (numerical)`, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Adding the equation and R-squared value
  stat_poly_eq(formula = y ~ x,
               aes(label = paste(after_stat(eq.label), 
                                 after_stat(rr.label), 
                                 sep = "~~~")),
               parse = TRUE) +
   labs(x = "Depth",
       y =  "C (g/kg)") +
  theme_test()

```



## 6.3. Al-ox

```{r}
library(ggpmisc)

meta %>% 
  ggplot(aes(x = `Al-ox`, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Adding the equation and R-squared value
  stat_poly_eq(formula = y ~ x,
               aes(label = paste(after_stat(eq.label), 
                                 after_stat(rr.label), 
                                 sep = "~~~")),
               parse = TRUE) +
   labs(x = "Al-ox (mmol/kg)",
       y =  "C (g/kg)") + 
  theme_test()


```




## 6.3. CAZyme_PCoA1

```{r}
library(ggpmisc)

df %>% 
 ggplot(aes(x = CAZyme_PCoA1, y = C)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Adding the equation and R-squared value
  stat_poly_eq(formula = y ~ x,
               aes(label = paste(after_stat(eq.label), 
                                 after_stat(rr.label), 
                                 sep = "~~~")),
               parse = TRUE) +
   labs(x = "CAZyme PCoA1",
       y =  "C (g/kg)") +
  theme_test()

```






```{r}
rm(list = ls())
```




