---
title: "singleM"
author: "Laura_Hakkinen"
date: "2025-05-28"
output: html_document
---


# singleM analysis of raw reads

We will assess the prokaryotic diversity, Bray-Curtis based beta-diversity, and composition of prokaryotic phyla


For alpha- and beta-diversity measures, we are using the S3.5.ribosomal_protein_S2_rpsB gene based phylotypes with 97% identity.

# 1. Alpha-diversity


## 1.1. Load data

```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)

setwd('C:\\Users\\lehakkin\\RStudio_computer')

# load the rpsB singleM data phyloseq object
load(file = 'ps_clustered_97')#ps

meta <- meta(ps)

```


## 1.2. Calculate diversity

This is how we would calculate diversity, but as we have it in the data already let's not run it now

```{r}
#div <- microbiome::alpha(ps, index = "all")
#div <- as.data.frame(div)
#div$sampleID <- row.names(div)
#meta <- left_join(meta, div, by = "sampleID")
#meta <- as.data.frame(meta)
#row.names(meta) <- meta$sampleID

# Assign the estimatedmeta# Assign the estimated diversity to sample metadata
#sample_data(ps) <- sample_data(meta)


#meta <- meta(ps)
```





## 1.3. Plot diversity



```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


### 1.3.1. Chao1

```{r}

# chao
chao <- meta %>%
  group_by(sample_type, depth_numerical) %>%
  summarise(mean = mean(chao1, na.rm = TRUE), se = (sd(chao1, na.rm = TRUE)/sqrt(length((chao1))))) %>%
  ungroup() %>%
  ggplot(aes(y=mean, depth_numerical, color=sample_type)) +
  geom_line(linetype = "dashed") +
  geom_point(size=3, position=position_dodge(1.2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5, position=position_dodge(1.2)) + coord_flip() + scale_x_reverse() + theme(plot.title = element_text(size = 20, hjust = 0.5)) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=16),
        axis.title = element_text(size=16),
        legend.text = element_text(size=16),
        legend.title = element_text(size=16),
        title = element_text(size=18)) +
  scale_y_continuous(name="Prokaryotic chao1") +
  labs(x = "depth") + theme(legend.title = element_blank()) + scale_colour_manual(values = MyPalette)

chao <- chao + scale_x_reverse(breaks = c(60, 35, 25, 15, 5), labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")) + theme(axis.title.y = element_blank())

print(chao)

```





### 1.3.2. Shannon

```{r}
# shannon
shannon <- meta %>%
  dplyr::group_by(sample_type, depth_numerical) %>%
  dplyr::summarise(mean = mean(shannon, na.rm = TRUE), se = (sd(shannon, na.rm = TRUE)/sqrt(length((shannon))))) %>%
  dplyr::ungroup() %>%
  ggplot(aes(y=mean, depth_numerical, color=sample_type)) +
  geom_line(linetype = "dashed") +
  geom_point(size=3, position=position_dodge(1.2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5, position=position_dodge(1.2)) + coord_flip() + scale_x_reverse() + theme(plot.title = element_text(size = 20, hjust = 0.5)) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=16),
        axis.title = element_text(size=16),
        legend.text = element_text(size=18),
        legend.title = element_text(size=16),
        title = element_text(size=18)) +
  scale_y_continuous(name="Prokaryotic Shannon") +
  labs(x = "depth") + theme(legend.title = element_blank()) + scale_colour_manual(values = MyPalette)

shannon <- shannon + scale_x_reverse(breaks = c(60, 35, 25, 15, 5), labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm"))+ theme(axis.title.y = element_blank())

shannon
```

### 1.3.3. Richness

```{r}

# richness
rich <- meta %>%
  group_by(sample_type, depth_numerical) %>%
  summarise(mean = mean(observed, na.rm = TRUE), se = (sd(observed, na.rm = TRUE)/sqrt(length((observed))))) %>%
  ungroup() %>%
  ggplot(aes(y=mean, depth_numerical, color=sample_type)) +
  geom_line(linetype = "dashed") +
  geom_point(size=3, position=position_dodge(1.2)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.5, position=position_dodge(1.2)) + coord_flip() + scale_x_reverse() + theme(plot.title = element_text(size = 20, hjust = 0.5)) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=16),
        axis.title = element_text(size=16),
        legend.text = element_text(size=16),
        legend.title = element_text(size=16),
        title = element_text(size=18)) +
  scale_y_continuous(name="Prokaryotic richness") +
  labs(x = "depth") + theme(legend.title = element_blank()) + scale_colour_manual(values = MyPalette)

rich <- rich + scale_x_reverse(breaks = c(60, 35, 25, 15, 5), labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")) + theme(axis.title.y = element_blank())

print(rich)

```




# 1.4. Test alpha-diversity

GENERALIZED MIXED MODEL

Let's focus on Chao1



I dont' need to use or I shouldn't use poisson, which is meant for count data, or nbinom2 as Chao1 values are no longer integer values!




```{r}
meta <- subset(meta, sample_type!="forest")

meta$treatment <- factor(meta$sample_type, 
                            levels = c("meadow", "organic", "conventional"))

levels(meta$treatment)

meta$depth <- factor(meta$depth, 
                            levels = c("0...10", "10...20", "20...30", "30...40", "40..."))

levels(meta$depth)


```

# 1. Data distribution

```{r}
# Basic density plot in ggplot2
ggplot(meta, aes(x = chao1)) +
  geom_density()
```




```{r}
shapiro.test(meta$chao1)
```





# 2. GLM




```{r}
mx <- glmmTMB(chao1 ~ depth * treatment + (1|plot), 
              data = meta, 
              tweedie(link = "log"),
              dispformula = ~ depth,  # Variance varies by depth!
              control = glmmTMBControl(
                optCtrl = list(iter.max = 1000, eval.max = 1000)
              ))

# 2. Check model diagnostics
summary(mx)
Anova(mx, type = "II", test = "Chisq")  # Type II tests

# 3. Validate residuals
sim_res <- simulateResiduals(mx, n = 250)
plot(sim_res)
testResiduals(sim_res)

# 1. Test if you need it:
mx0 <- glmmTMB(chao1 ~ depth * treatment + (1|plot), 
              data = meta, 
              tweedie(link = "log"),
              control = glmmTMBControl(
                optCtrl = list(iter.max = 1000, eval.max = 1000)
              ))

# Likelihood ratio test
anova(mx0, mx)  # If p < 0.05, include dispformula
```


# 3. Final model: m1


```{r}
# 1. Fit model (with singular fit - this is fine!)
m1 <- glmmTMB(chao1 ~ depth * treatment + (1|plot), 
              data = meta, 
              tweedie(link = "log"),
              control = glmmTMBControl(
                optCtrl = list(iter.max = 1000, eval.max = 1000)
              ))

# 2. Check model diagnostics
summary(m1)
Anova(m1, type = "II", test = "Chisq")  # Type II tests

##########################
logLik(m1)
##########################
# 1. Get predictions using ONLY the fixed effects (depth * treatment)
# re.form = NA drops the random effects from the prediction
fixed_preds <- predict(m1, type = "response", re.form = NA)

# 2. Get predictions using BOTH fixed and random effects
full_preds <- predict(m1, type = "response")

# 3. Get your actual raw response data from the model
raw_data <- m1$frame$chao1

# 4. Compute manual R2 values
R2_marginal_manual    <- var(fixed_preds) / var(raw_data)
R2_conditional_manual <- var(full_preds) / var(raw_data)

# Print results
print(R2_marginal_manual)
print(R2_conditional_manual)

#########################

# 3. Validate residuals
sim_res <- simulateResiduals(m1, n = 250)
plot(sim_res)
testResiduals(sim_res)


##########################

# Force emmeans to always calculate back-transformed real-scale proportions (0 to 1)
# while performing pairwise tests safely on the internal logit scale.

library(emmeans)
library(multcomp)

# =========================================================================
# 1. INTERACTION POST-HOC 
# =========================================================================

# Option A: Comparing treatments within each specific depth
emm_inter_A <- emmeans(m1, ~ treatment | depth, type = "response")
# Get Letters:
cld_interaction_A <- as.data.frame(cld(emm_inter_A, adjust = "tukey", Letters = letters, reversed = TRUE))
# Get P-values:
p_interaction_A <- as.data.frame(contrast(emm_inter_A, method = "pairwise", adjust = "tukey"))


# Option B: Comparing depths within each specific treatment
emm_inter_B <- emmeans(m1, ~ depth | treatment, type = "response")
# Get Letters:
cld_interaction_B <- as.data.frame(cld(emm_inter_B, adjust = "tukey", Letters = letters, reversed = TRUE))
# Get P-values:
p_interaction_B <- as.data.frame(contrast(emm_inter_B, method = "pairwise", adjust = "tukey"))


# =========================================================================
# 2. DEPTH MAIN EFFECT POST-HOC 
# =========================================================================

emm_depth_main <- emmeans(m1, ~ depth, type = "response")
# Get Letters:
cld_depth_main <- as.data.frame(cld(emm_depth_main, adjust = "tukey", Letters = letters, reversed = TRUE))
# Get P-values:
p_depth_main <- as.data.frame(contrast(emm_depth_main, method = "pairwise", adjust = "tukey"))


# =========================================================================
# 3. TREATMENT MAIN EFFECT POST-HOC 
# =========================================================================

emm_treatment_main <- emmeans(m1, ~ treatment, type = "response")
# Get Letters:
cld_treatment_main <- as.data.frame(cld(emm_treatment_main, adjust = "tukey", Letters = letters, reversed = TRUE))
# Get P-values:
p_treatment_main <- as.data.frame(contrast(emm_treatment_main, method = "pairwise", adjust = "tukey"))


# =========================================================================
# 4. Print all outputs to your console
# =========================================================================
print("--- Interaction: Treatment within Depth (LETTERS & P-VALUES) ---")
print(cld_interaction_A)
print(p_interaction_A)

print("--- Interaction: Depth within Treatment (LETTERS & P-VALUES) ---")
print(cld_interaction_B)
print(p_interaction_B)

print("--- Main Effect: Depth (LETTERS & P-VALUES) ---")
print(cld_depth_main)
print(p_depth_main)

print("--- Main Effect: Treatment (LETTERS & P-VALUES) ---")
print(cld_treatment_main)
print(p_treatment_main)


```





# 2. Beta-diversity


```{r}
ps_RA <- microbiome::transform(ps, "compositional")
```


Let's first quickly see how the ordination looks. Later, let's use better color options.

```{r}
OTU = as(otu_table(ps_RA), "matrix")
# transpose if necessary
if(taxa_are_rows(ps_RA)){OTU <- t(OTU)}
# Coerce to data.frame
OTU = as.data.frame(OTU)
OTU <- as.matrix(OTU)

bray_dist <- vegan::vegdist(OTU, method="bray")
str(bray_dist)

# use k = 3 so we'll get calculations for three axes
pcoa <- cmdscale(bray_dist, eig=TRUE, k = 3)
ordiplot(scores(pcoa),display="sites", type="points")
```

## 2.1. PCoA

Let's first make PCoA ordination with axes 1 and 2.

### 2.1.1. Env. variables

```{r}
# Post-Hoc Projections of environmental variables
# envfit in vegan projects points onto vectors that have maximum correlation with corresponding environmental variables
pcoa.env12 <- envfit(pcoa, meta[ ,c("pH_H2O", "EC_uScm", "C_g_per_kg", "N_gkg", "TP_gkg", "sample_type", "depth_numerical", "DOC_mgkg", "Pinorg_mgkg", "Porg_mgkg", "log_root", "C_per_N", "Feox_mmolkg", "Alox_mmolkg", "oxides_mmolkg", "PH2O_mgkg")], na.rm = TRUE, choices=c(1:2), permutations = 999)


# main effects
pcoa.env12

ef12.adj <- pcoa.env12
pvals.adj <- p.adjust (pcoa.env12$vectors$pvals, method = 'bonferroni')
ef12.adj$vectors$pvals <- pvals.adj
ef12.adj

```


To plot (classical) mds (which is equivalent to PCoA) with ggplot a new datasheet needs to be created which contains the x,y points for each site. You can do this by calling the scores of you mds.


```{r}
site.scrs <- as.data.frame(scores(pcoa, display = "sites")) #save pcoa results into dataframe
site.scrs <- cbind(site.scrs, soil_type = meta$sample_type) #add grouping variable "soil_type" to dataframe
site.scrs <- cbind(site.scrs, depth = meta$depth) #add grouping variable of depth

head(site.scrs)
```


I will be following somewhat this tutorial for fitting the environmental variables etc:

[jkzorz github](https://jkzorz.github.io/2020/04/04/NMDS-extras.html)

To show environmental extrinsic variables another datasheet needs to be created

Citation from the [jkzorz github](https://jkzorz.github.io/2020/04/04/NMDS-extras.html)

*"Extracting the required information from the envfit result is a bit more complicated. The envfit output contains information on the length of the segments for each variable. The segments are scaled to the r2 value, so that the environmental variables with a longer segment are more strongly correlated with the data than those with a shorter segment. You can extract this information with scores. Then these lengths are further scaled to fit the plot. This is done with a multiplier that is analysis specific, and can be accessed using the command ordiArrowMul(en). Below I multiply the scores by this multiplier to keep the coordinates in the correct proportion."*

Because my data contained continuous and categorical environmental variables, I am extracting the information from both separately using the vectors and factors options respectively.

### 2.1.2 Categorial and continuous variables

```{r}

# first categorial variables
env.scores_cat12 <- as.data.frame(scores(pcoa.env12, display = "factors"))*0.25 # multiply the envfit scores to keep them in the frame of the ordination
env.scores_cat12 <- cbind(env.scores_cat12, env.variables = rownames(env.scores_cat12)) #and then gives them their names

env.scores_cat12 <- cbind(env.scores_cat12, pval = pcoa.env12$factors$pvals) # add pvalues to dataframe
sig.env.scores_cat12 <- subset(env.scores_cat12, pval<=0.05) #subset data to show variables significant at 0.05
sig.env.scores_cat12
# all were significant


# then continous variables
env.scores_cont12 <- as.data.frame(scores(pcoa.env12, display = "vectors"))*0.25 # multiply the envfit scores to keep them in the frame of the ordination
env.scores_cont12 <- cbind(env.scores_cont12, env.variables = rownames(env.scores_cont12)) #and then gives them their names

env.scores_cont12 <- cbind(env.scores_cont12, pval = pcoa.env12$vectors$pvals) # add pvalues to dataframe
sig.env.scores_cont12 <- subset(env.scores_cont12, pval<=0.05) #subset data to show variables significant at 0.05
sig.env.scores_cont12
# all were significant
```



Now we have the relevant information for plotting the ordination in ggplot


```{r}
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")

site.scrs$new_depth <- site.scrs$depth

site.scrs$new_depth <- gsub("...", "-", site.scrs$new_depth, fixed = TRUE)

site.scrs$new_depth[site.scrs$new_depth=="40-"] <- "40-80"

site.scrs$new_depth[site.scrs$new_depth=="0-10"] <- "0-10 cm"
site.scrs$new_depth[site.scrs$new_depth=="10-20"] <- "10-20 cm"
site.scrs$new_depth[site.scrs$new_depth=="20-30"] <- "20-30 cm"
site.scrs$new_depth[site.scrs$new_depth=="30-40"] <- "30-40 cm"
site.scrs$new_depth[site.scrs$new_depth=="40-80"] <- "40-80 cm"
```


The phyloseq package PCoA gives the axis %

```{r}
#First lets run the PCoA with phyloseq to get the axis percentages (plot_ordination gives axis percentages)

GP.ord <- ordinate(ps_RA, "PCoA", "bray")
p2 = plot_ordination(ps_RA, GP.ord, type="samples", color="sample_type", shape="depth")
p2
```

Remember to change the axis percentages accordingly below!!!



### 2.1.3. Plot PCoA


```{r}
pcoa.plot <- ggplot() + geom_point(data=site.scrs, aes(Dim1, Dim2, colour = factor(site.scrs$soil_type), shape = factor(site.scrs$new_depth)), size = 6, alpha = 0.6, stroke = 1.5) + theme_cowplot() + theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid")) + labs(colour = "", shape = "") + theme(legend.text = element_text(size = 12), axis.text = element_text(size = 16)) + scale_colour_manual(values = MyPalette) + labs(y = "PCoA2 (4.1%)", x = "PCoA1 (11.2%)")


pcoa.plot

```


### 2.1.4. Add env. variables to the PCoA

First, simplify the names


```{r}
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "pH_H2O"] <- "pH"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "C_g_per_kg"] <- "C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "N_gkg"] <- "N"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "TP_gkg"] <- "P-tot"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "depth_numerical"] <- "Depth"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "DOC_mgkg"] <- "DOC"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Porg_mgkg"] <- "P-org"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "log_root"] <- "log root"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "C_per_N"] <- "C/N"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Feox_mmolkg"] <- "Fe-ox"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Alox_mmolkg"] <- "Al-ox"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "depth"] <- "Depth"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "log root"] <- "Root biomass"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "S_mg_per_kg"] <- "S"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Mn_mg_per_kg"] <- "Mn"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "PH2O_mgkg"] <- "P-H2O"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "POM_C_percent"] <- "POM-C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "MAOM_C_percent"] <- "MAOM-C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "oxides_mmolkg"] <- "Oxides"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "EC_uScm"] <- "EC"
```



...then plot





```{r fig.dim = c(12, 12)}
p2 <- pcoa.plot +
  geom_segment(aes(x = 0, y = 0, xend = Dim1, yend = Dim2), data = sig.env.scores_cont12, size =1, alpha = 0.5, colour = "grey30") + ggrepel::geom_text_repel(data = sig.env.scores_cont12, aes(x = Dim1, y = Dim2), colour = "blue", fontface = "bold", label = sig.env.scores_cont12$env.variables, segment.size = 0.2, box.padding = unit(0.1, "lines"), point.padding = (0.1), force = 1, max.time = 30, nudge_y = 0.00, nudge_x = 0.00) + theme(legend.position = c(0.05, 0.3), legend.text = element_text(size = 12))

p2
```



## 2.2. PERMANOVA


No forest!


```{r}

# subset samples
ps_RA <- subset_samples(ps_RA, sample_type != "forest")

meta <- subset(meta, sample_type != "forest")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_RA),
              MARGIN = ifelse(taxa_are_rows(ps_RA), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_RA = prune_taxa((prev0 > 0), ps_RA)
ps_RA
```


### 2.2.1. Calculate Bray-Curtis (dis)similarities

```{r}
ps_RA_bray <- phyloseq::distance(ps_RA, method = "bray")
```


### 2.2.2. Homogenity of variances

Check that variance homogeneity assumptions hold (to ensure the reliability of the results). If  groups have signif. different spreads the permanova result may be potentially explained by that, rtaher than the groups.

Betadisper first calculates the average distance of group members to the group centroid in multivariate space (generated by a distance matrix). Then, an ANOVA is done to test if the dispersions (variances) of groups are different.



for management type

```{r}
anova(betadisper(ps_RA_bray, meta$sample_type))
```
         

We see that the ANOVA p-value is not significant meaning that the **homogeneity of variance assumption is met**



for depth

```{r}
anova(betadisper(ps_RA_bray, meta$depth))
```

We see that the ANOVA p-value is  significant meaning that **homogeneity of variance assumption is *NOT* met**

lets do post hoc analysis with Tukey's test to see which groups differ in relation to their variances

```{r}
TukeyHSD(betadisper(ps_RA_bray, meta$depth))
```

Dispersions differ significantly between 30-40 and the ones above it

It is not a problem, because I am not interested of comparing layers.


### 2.2.3. Run initial PERMANOVAs

First, I will do PERMANOVA so that I include all management types (later without forest)


```{r}
# first with just soil type and strata option
adonis2(formula = ps_RA_bray~ sample_type, data = meta, permutations = 9999, method = "bray", by = "margin", strata = meta$depth)
```

Significant




Depth

```{r}
# then with just depth and strata option
adonis2(formula = ps_RA_bray~ depth, data = meta, permutations = 9999, method = "bray", by = "margin", strata = meta$sample_type)
```

Depth has a larger effect than treatment (sample_type). So, let's put it first in the model

### 2.2.4. Run final PERMANOVA

Depth*Treatment

For the full model it matters which "by" option we choose. When by="terms" significance for each term is calculated sequentially from first to last, so that the order of terms matter. We will use this, because with sequential analysis we will get R2 values that sum up to 1, and will also get the significance and R2 values for each interaction term separately rather than for the interaction alone.


```{r}
final <- adonis2(formula = ps_RA_bray ~ depth * sample_type, data = meta, permutations = 9999, method = "bray", by = "terms")
final
```

Interaction is significant













# 3. Taxonomic composition

For the phyla composition; we will be using the singleM data with all the single-copy marker genes (results from singleM profile pipeline), as this gives a better taxonomic assignment than just one gene (rpsB gene above)

First let's clear the environment and load the data with all the genes

```{r}
rm(list = ls())
```


```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)

setwd('C:\\Users\\lehakkin\\RStudio_computer')

load(file='ps_profile_Phyla')

meta <- meta(ps)
```





I need to remove the unassigned samples, they are non-bacterial seqs and could not be identified well enough with the singleM pipeline

```{r}
ps <- subset_taxa(ps, domain!="unassigned")

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps),
              MARGIN = ifelse(taxa_are_rows(ps), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps = prune_taxa((prev0 > 0), ps)
```


## 3.1. Phyla composition barplot

```{r fig.dim = c(10, 5)}
ps_RA <- transform(ps, "compositional")
ps_RA <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 3/140, prevalence = 3/140)

#create data table
ps_RA_phyla_df <-  psmelt(ps_RA)

ps_RA_phyla_df$sample_type <- factor(ps_RA_phyla_df$sample_type, levels = c("forest", "meadow", "organic", "conventional"))



#Create color palette
cbbPalette <- c("#b2b2b2",  "#8DD3C7",  "#FFED6F", "#239FD7",  "#F0D1E1",  "#FF6199", "#FF8DB5", "#C191C2",  "#BCD868",  "#FFFFC6", "#D0D9CD",  "#03B7EF", "#C66542",  "#8BC081",  "#EE9A04",  "deeppink", "#F68378",  "#8D6942", "#8DD3C7", "#F3B962", "#BCD868", "#6E99BE", "#F0D1E1", "#C191C2", "#FFFFC6", "darkgreen", "#D0D9CD","#FF8DB5", "#8BC081",  "#b71c1c", "deeppink", "#d6849a", "darkolivegreen1")


sampletype_names <- list(
  'forest' = "forest", 'meadow' = "meadow", 'organic' = "organic", 'conventional' = "conventional"
)



# check unique values for phylum
unique(ps_RA_phyla_df$phylum)

#ps_RA_phyla_df$phylum[ps_RA_phyla_df$phylum=="unassigned"] <- "Unassigned"

ps_RA_phyla_df$phylum <- as.factor(ps_RA_phyla_df$phylum)
levels(ps_RA_phyla_df$phylum)

ps_RA_phyla_df$phylum <- factor(ps_RA_phyla_df$phylum, levels = c("Other", "Acidobacteriota", "Actinomycetota", "Bacteroidota", "Chloroflexota", "Desulfobacterota", "Desulfobacterota_B", "Gemmatimonadota", "Methylomirabilota", "Myxococcota", "Nitrospirota", "Patescibacteriota", "Planctomycetota", "Pseudomonadota", "Thermoproteota", "Verrucomicrobiota"))

levels(ps_RA_phyla_df$phylum)

# Make new depth variable
ps_RA_phyla_df$Depth <- ps_RA_phyla_df$depth

ps_RA_phyla_df$Depth <- gsub("...", "-", ps_RA_phyla_df$Depth, fixed = TRUE)

ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="40-"] <- "40-80"


ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="0-10"] <- "0-10 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="10-20"] <- "10-20 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="20-30"] <- "20-30 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="30-40"] <- "30-40 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="40-80"] <- "40-80 cm"

#Create a plot
phylumF  <- ggplot(ps_RA_phyla_df, aes(x = Depth ,y = Abundance, fill = phylum)) + geom_bar(stat="identity", position="fill") + scale_fill_manual(values = cbbPalette) + facet_grid (cols = vars(sample_type)) + geom_bar(stat="identity", position="fill") + scale_fill_manual(values = cbbPalette) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.text = element_text(size=16),
        legend.title = element_text(size=18),
        legend.spacing.y = unit(0, 'cm'),
        legend.key.size = unit(0.8, 'cm'),
        title = element_text(size=18))+ theme(axis.text.x = element_text(angle = 45, hjust = 1)) + guides(fill = guide_legend(title="")) + theme(strip.text.x = element_text(size = 22)) +
  ylab(label = "Relative abundance") + theme(legend.position="right") +  xlab("Soil layer")

phylumF
```



## 3.2. Most common phyla

```{r}

ps_RA <- transform(ps, "compositional")
ps_RA <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 0/140, prevalence = 0/140)

#create data table
ps_RA_phyla_df <-  psmelt(ps_RA)

ps_RA_phyla_df$sample_type <- factor(ps_RA_phyla_df$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

MySummary <- ps_RA_phyla_df %>%
  group_by(phylum, depth) %>%
  summarize(mean_abund = mean(Abundance, na.rm=TRUE)) 
MySummary
```

same without forest

```{r}
ps.x <- subset_samples(ps, sample_type!="forest")
ps.x_RA <- transform(ps.x, "compositional")
ps.x_RA <- microbiome::aggregate_rare(ps.x_RA, level = 'phylum', detection = 0/140, prevalence = 0/140)


#create data table
ps_RA_phyla_df <-  psmelt(ps.x_RA)

ps_RA_phyla_df$sample_type <- factor(ps_RA_phyla_df$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

MySummary <- ps_RA_phyla_df %>%
  group_by(phylum, sample_type) %>%
  summarize(mean_abund = mean(Abundance, na.rm=TRUE)) 
MySummary
```




## 3.3. Differential abundance analysis: MaAsLin2

I need to have the phylum name in the taxa names in the phyloseq object  (now it is just phylum 1, phylum2, etc)

```{r}
tax <- as.data.frame(tax_table(ps))

tax$name <- row.names(tax)
x <- tax[, c(2,3)]

otu <- as.data.frame(otu_table(ps))
otu$name <- row.names(otu)

otu <- left_join(otu, x, by="name")
row.names(otu) <- otu$phylum
otu <- as.data.frame(otu)
otu <- subset(otu, select = -c(name, phylum))


row.names(tax) <- NULL
row.names(tax) <- tax$phylum
tax <- subset(tax, select = -c(name))

meta <- meta(ps)

# make all the elements, OTU, TAX and samples, into phyloseq compatible format
samples <- sample_data(meta)
OTU <- as.matrix(otu)
OTU <- otu_table(OTU, taxa_are_rows = TRUE)
tax <- tax_table(as.matrix(tax))

# finally combine all the elemnt to make a phyloseq object
ps <- phyloseq(OTU, tax, samples)
```





MaAsLin2 (Microbiome Multivariate Analysis)
Handles relative abundance data.

Works with generalized linear models (GLMs).


I will exclude the forest.



```{r}
ps_RA <- transform(ps, "compositional")
ps_RA
  
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA, sample_type!="forest")
#This will return a ps object with the samples removed

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)

```

### 3.3.1. with Meadow

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\READ_MaAsLin2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

# This is how the analysis would be run:

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "maaslin_RA_CLR_PHYLA_NoForest_READ_BASED_Phyla",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # no 
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM",
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )

```





### 3.3.2. MaAsLin2 without meadow

We will also run without the meadow to be able to compare the organic and conventional to each other

```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)

```




```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\READ_MaAsLin2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "maaslin_RA_CLR_PHYLA_NoForest_READ_BASED_Phyla_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # no 
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM",
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional'
#  )

```




```{r}
rm(list = ls())
```







