---
title: "Untitled"
output: html_document
date: "2025-12-17"
---



# Phyla encoding KEGG and CAZyme genes

Let's see which taxa (phyla) are encoding the KEGG and CAZyme genes and if they are differently represesnted


# 1. KEGG

## 1.1. Load KEGG data



```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)


setwd('C:\\Users\\lehakkin\\RStudio_computer')

load(file = 'KOfam_PHYLUM_ps')
ps
meta <- meta(ps)
```




## 2.1. KEGG phyla composition bar plot

```{r fig.dim = c(10, 5)}
ps_RA <- transform(ps, "compositional")
tax <- as.data.frame(tax_table(ps))

# only keep "phyla" and "phyla+geneID" columns
tax <- tax[, c(6, 10)]

tax_table(ps_RA) <- tax_table(as.matrix(tax))

#ps_RA <- transform(ps, "compositional")
ps_RA_p <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 1/100, prevalence = 2/140)
ps_RA_p


#create data table
ps_RA_phyla_df <-  psmelt(ps_RA_p)

#Create color palette
cbbPalette <- c("#b2b2b2", "#8DD3C7", "#FFED6F", "#CAAEC5", "#F68378", "#8D6942", "#F3B962", "#BCD868", "#6E99BE", "#F0D1E1", "#C191C2", "#FFFFC6", "darkgreen", "#D0D9CD", "#8BC081", "#FF8DB5", "#b71c1c", "deeppink", "#d6849a", "darkolivegreen1")


sampletype_names <- list(
  'forest' = "forest", 'meadow' = "meadow", 'organic' = "organic", 'conventional' = "conventional"
)

sampletype_labeller <- function(variable,value){
  return(sampletype_names[value])
}


# check unique values for phylum
unique(ps_RA_phyla_df$phylum)



ps_RA_phyla_df$phylum <- as.factor(ps_RA_phyla_df$phylum)
levels(ps_RA_phyla_df$phylum)
ps_RA_phyla_df$phylum <- factor(ps_RA_phyla_df$phylum, levels = c("Other", "Acidobacteriota", "Actinomycetota", "Bacillota", "Bacteria", "Candidatus Levyibacteriota", "Candidatus Methylomirabilota", "Candidatus Rokuibacteriota", "Candidatus Sysuimicrobiota", "Chloroflexota", "Gemmatimonadota",  "Myxococcota", "Nitrososphaerota", "Nitrospirota", "Pseudomonadota", "Thermodesulfobacteriota", "Thermomicrobiota", "Verrucomicrobiota"))

levels(ps_RA_phyla_df$phylum)

# Make new depth variable
ps_RA_phyla_df$Depth <- ps_RA_phyla_df$depth

ps_RA_phyla_df$Depth <- gsub("...", "-", ps_RA_phyla_df$Depth, fixed = TRUE)

ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="40-"] <- "40-80"


ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="0-10"] <- "0-10 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="10-20"] <- "10-20 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="20-30"] <- "20-30 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="30-40"] <- "30-40 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="40-80"] <- "40-80 cm"

#Create a plot
phylumF  <- ggplot(ps_RA_phyla_df, aes(x = Depth ,y = Abundance, fill = phylum)) + geom_bar(stat="identity", position="fill") + scale_fill_manual(values = cbbPalette) + facet_grid (cols = vars(sample_type)) + geom_bar(stat="identity", position="fill") + scale_fill_manual(values = cbbPalette) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.text = element_text(size=16),
        legend.title = element_text(size=18),
        legend.spacing.y = unit(0, 'cm'),
        legend.key.size = unit(0.8, 'cm'),
        title = element_text(size=18))+ theme(axis.text.x = element_text(angle = 45, hjust = 1)) + guides(fill = guide_legend(title="")) + theme(strip.text.x = element_text(size = 22)) +
  ylab(label = "Relative abundance") + theme(legend.position="right") +  xlab("Soil layer")

phylumF
```


## 2.2. numerical data


```{r}
ps_RA <- transform(ps, "compositional")

ps_RA_p <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 0/100, prevalence = 0/140)
ps_RA_p


#create data table
ps_RA_phyla_df <-  psmelt(ps_RA_p)
```




```{r}
ps_RA_phyla_df$percentage <- ps_RA_phyla_df$Abundance * 100

KEGG <- ps_RA_phyla_df %>%
  group_by(sample_type, depth, phylum) %>%
  summarize(mean_abund = mean(percentage, na.rm=TRUE)) 
head(KEGG)
```



## 2.3. DIfferential abundance analysis


### 2.3.1. With Meadow


```{r}
ps_RA <- transform(ps, "compositional")
tax <- as.data.frame(tax_table(ps))

tax <- tax[, c(6, 10)]

tax_table(ps_RA) <- tax_table(as.matrix(tax))

ps_RA_p <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 1/100, prevalence = 2/140)
ps_RA_p

```

```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_p, sample_type!="forest")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\Gene_PHYLA_MaAsLin2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "KEGG_TREATMENT_and_LAYER",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )
```


### 2.3.2. WO Meadow


```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_p, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\Gene_PHYLA_MaAsLin2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "KEGG_TREATMENT_and_LAYER_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional'
#  )
```


# 2. CAZyme

rename KEGG data

```{r}
# lets first save KEGG data with new name
ps_RA <- transform(ps, "compositional")
tax <- as.data.frame(tax_table(ps))

tax <- tax[, c(6, 10)]

tax_table(ps_RA) <- tax_table(as.matrix(tax))

ps_RA_p_KEGG <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 0/100, prevalence = 0/140)
ps_RA_p_KEGG

```


Remove forest from the analysis

```{r}
#To remove those from your phyloseq object
ps_RA_p_KEGG <- subset_samples(ps_RA_p_KEGG, sample_type!="forest")
#This will return a ps object with the samples removed
ps_RA_p_KEGG

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_RA_p_KEGG),
              MARGIN = ifelse(taxa_are_rows(ps_RA_p_KEGG), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_RA_p_KEGG = prune_taxa((prev0 > 0), ps_RA_p_KEGG)
ps_RA_p_KEGG

```


## 2.1. Load CAZyme data

```{r}
setwd('C:\\Users\\lehakkin\\RStudio_computer')
load(file = 'CAZyme_PHYLUM_ps')
meta <- meta(ps)
```



```{r fig.dim = c(10, 5)}
tax <- as.data.frame(tax_table(ps))

tax <- tax[, -c(1, 2, 7)]

ps_RA <- transform(ps, "compositional")
tax_table(ps_RA) <- tax_table(as.matrix(tax))

#ps_RA <- transform(ps, "compositional")
ps_RA_p <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 1/100, prevalence = 2/140)
ps_RA_p


#create data table
ps_RA_phyla_df <-  psmelt(ps_RA_p)

#Create color palette
cbbPalette <- c("#b2b2b2", "#8DD3C7", "#FFED6F", "#CAAEC5", "#F68378", "#239FD7", "#8D6942", "#F3B962", "#BCD868", "#6E99BE", "#F0D1E1", "#FE4C48", "#C191C2", "darkolivegreen1","#FFFFC6","darkgreen", "#D0D9CD","#C66542", "#8BC081", "#FF8DB5", "#b71c1c", "deeppink", "#d6849a",  "#8D6942", "#F3B962", "#BCD868", "#6E99BE","#FF8DB5", "#b71c1c", "#EB730F")




sampletype_names <- list(
  'forest' = "forest", 'meadow' = "meadow", 'organic' = "organic", 'conventional' = "conventional"
)

sampletype_labeller <- function(variable,value){
  return(sampletype_names[value])
}


# check unique values for phylum
unique(ps_RA_phyla_df$phylum)


ps_RA_phyla_df$phylum <- as.factor(ps_RA_phyla_df$phylum)
levels(ps_RA_phyla_df$phylum)

ps_RA_phyla_df$phylum <- factor(ps_RA_phyla_df$phylum, levels = c("Other",  "Acidobacteriota", "Actinomycetota", "Bacillota", "Bacteria", "Bacteroidota","Candidatus Levyibacteriota", "Candidatus Methylomirabilota", "Candidatus Rokuibacteriota", "Candidatus Sysuimicrobiota", "Chloroflexota",  "Cyanobacteriota", "Gemmatimonadota", "Methanobacteriota", "Myxococcota","Nitrososphaerota", "Nitrospirota", "Planctomycetota", "Pseudomonadota", "Thermodesulfobacteriota",  "Thermomicrobiota", "Verrucomicrobiota"))

levels(ps_RA_phyla_df$phylum)

# Make new depth variable
ps_RA_phyla_df$Depth <- ps_RA_phyla_df$depth

ps_RA_phyla_df$Depth <- gsub("...", "-", ps_RA_phyla_df$Depth, fixed = TRUE)

ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="40-"] <- "40-80"


ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="0-10"] <- "0-10 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="10-20"] <- "10-20 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="20-30"] <- "20-30 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="30-40"] <- "30-40 cm"
ps_RA_phyla_df$Depth[ps_RA_phyla_df$Depth=="40-80"] <- "40-80 cm"

#Create a plot
phylumCAZYme  <- ggplot(ps_RA_phyla_df, aes(x = Depth ,y = Abundance, fill = phylum)) + geom_bar(stat="identity", position="fill") + scale_fill_manual(values = cbbPalette) + facet_grid (cols = vars(sample_type)) + geom_bar(stat="identity", position="fill") + scale_fill_manual(values = cbbPalette) + theme(panel.border = element_rect(colour = "black", fill=NA, size=0.5)) + theme_cowplot() +
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.text = element_text(size=14),
        legend.title = element_text(size=18),
        legend.spacing.y = unit(0, 'cm'),
        legend.key.size = unit(0.7, 'cm'),
        title = element_text(size=18))+ theme(axis.text.x = element_text(angle = 45, hjust = 1)) + guides(fill = guide_legend(title="", ncol = 1)) + theme(strip.text.x = element_text(size = 22)) +
  ylab(label = "Relative abundance") + theme(legend.position="right") +  xlab("Soil layer")

phylumCAZYme
```

## 2.2. numerical data


```{r}
ps_RA <- transform(ps, "compositional")

ps_RA_p <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 0/100, prevalence = 0/140)
ps_RA_p


#create data table
ps_RA_phyla_df <-  psmelt(ps_RA_p)
```




```{r}

ps_RA_phyla_df$percentage <- ps_RA_phyla_df$Abundance * 100

cazyme <- ps_RA_phyla_df %>%
  dplyr::group_by(sample_type, depth, phylum) %>%
  dplyr::summarize(mean_abund = mean(percentage, na.rm=TRUE)) 
head(cazyme)



```



```{r}

ps_RA_phyla_df$percentage <- ps_RA_phyla_df$Abundance * 100

cazyme <- ps_RA_phyla_df %>%
  dplyr::group_by(sample_type, phylum) %>%
  dplyr::summarize(mean_abund = mean(percentage, na.rm=TRUE)) 
head(cazyme)



```


```{r}

ps_RA_phyla_df$percentage <- ps_RA_phyla_df$Abundance * 100

cazyme <- ps_RA_phyla_df %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarize(mean_abund = mean(percentage, na.rm=TRUE)) 
head(cazyme)



```

## 2.3. cazyme class



```{r}
setwd('C:\\Users\\lehakkin\\RStudio_computer')
load(file = 'CAZyme_FAMILY_PHYLUM_ps_RA')
meta <- meta(ps_cazyme_phylum)
```


```{r}
df <- psmelt(ps_cazyme_phylum)

```




```{r}
summary <- df %>%
  group_by(phylum, cazyme) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
head(summary)
```

Get the phyla with the 5 highest values in a column called mean for every value (character) in column called cazyme

```{r}
summary_top5_phylum <- summary %>%
  group_by(cazyme) %>%
  slice_max(mean, n = 5, with_ties = FALSE) %>%
  dplyr::select(cazyme, phylum, mean, se) %>%
  ungroup()

summary_top5_phylum
```


most common CAZymes

```{r}
tax <- as.data.frame(tax_table(ps_cazyme_phylum))
tax <- tax[, c(9, 1:8, 10)]
tax_table(ps_cazyme_phylum) <- tax_table(as.matrix(tax))
ps_cazyme_phylum_CAZyme_family <- aggregate_rare(ps_cazyme_phylum, level = "cazyme", detection = 0, prevalence = 0)

df_x <- psmelt(ps_cazyme_phylum_CAZyme_family)

x <- df_x %>%
  group_by(cazyme) %>%
  summarise(mean = mean(Abundance, na.rm = TRUE), se = (sd(Abundance, na.rm = TRUE)/sqrt(length((Abundance)))))
x

```


## 2.4. Differential abundance analysis




### 2.4.1. With Meadow





```{r}
tax <- as.data.frame(tax_table(ps))

tax <- tax[, -c(1, 2, 7)]

tax_table(ps_RA) <- tax_table(as.matrix(tax))

ps_RA_p <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 1/100, prevalence = 2/140)
ps_RA_p

```

```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_p, sample_type!="forest")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\Gene_PHYLA_MaAsLin2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "CAZyme_TREATMENT_and_LAYER",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )
```


# 2. WO Meadow


```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_p, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\Gene_PHYLA_MaAsLin2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "CAZyme_TREATMENT_and_LAYER_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional'
#  )
```








# 3. Compare KEGG and CAZyme




```{r}
ps_RA <- transform(ps, "compositional")
tax <- as.data.frame(tax_table(ps))

tax <- tax[, -c(1, 2, 7)]

tax_table(ps_RA) <- tax_table(as.matrix(tax))

ps_RA_p_CAZyme <- microbiome::aggregate_rare(ps_RA, level = 'phylum', detection = 0/100, prevalence = 0/140)
ps_RA_p_CAZyme

```

remove forest

```{r}
#To remove those from your phyloseq object
ps_RA_p_CAZyme <- subset_samples(ps_RA_p_CAZyme, sample_type!="forest")
#This will return a ps object with the samples removed
ps_RA_p_CAZyme

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_RA_p_CAZyme),
              MARGIN = ifelse(taxa_are_rows(ps_RA_p_CAZyme), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_RA_p_CAZyme = prune_taxa((prev0 > 0), ps_RA_p_CAZyme)
ps_RA_p_CAZyme

```

## 3.2. combine KEGG and CAZyme data

```{r}
#create data table
KEGG_df <-  psmelt(ps_RA_p_KEGG)

colnames(KEGG_df)[colnames(KEGG_df) == 'Abundance'] <- 'KEGG'
KEGG_df <- KEGG_df[, c(1,2,3)]


CAZyme_df <-  psmelt(ps_RA_p_CAZyme)
colnames(CAZyme_df)[colnames(CAZyme_df) == 'Abundance'] <- 'CAZyme'
CAZyme_df <- CAZyme_df[, c(1,2,3)]

df <- full_join(KEGG_df, CAZyme_df, by = c("OTU", "Sample"))

# then get the metadata correctly
df_full <- left_join(df, meta, by = c("Sample"="sampleID"))

# pivot longer
df_full_long <- df_full %>% pivot_longer(cols=c('KEGG', 'CAZyme'),
                    names_to='database',
                    values_to='Abundance') 

# NAs to zeros
df_full_long <- df_full_long %>% mutate(Abundance = ifelse(is.na(Abundance), 0, Abundance))

# use percentages instead of proportion
df_full_long$Abundance <- df_full_long$Abundance*100
```





## 3.3. do the comparison


with BH p value adjustment

```{r}
library(dplyr)
library(car)
library(tidyr)

# Function to analyze one OTU
analyze_otu <- function(taxa, data) {
  # Filter data for current OTU
  df <- filter(data, OTU == taxa)
  
   # Convert database to factor explicitly
  df$database <- as.factor(df$database)
  
  # Levene's test
  levene_result <- tryCatch(
    {
      result <- leveneTest(Abundance ~ database, df)
      result$`Pr(>F)`[1]
    },
    error = function(e) NA_real_
  )
  
  # Kruskal-Wallis test
  kruskal_result <- tryCatch(
    {
      result <- kruskal.test(Abundance ~ database, data = df)
      result$p.value
    },
    error = function(e) NA_real_
  )
  
  # Calculate mean and SE for each database
  summary_df <- df %>%
    group_by(database) %>%
    summarise(
      mean = mean(Abundance, na.rm = TRUE),
      se = sd(Abundance, na.rm = TRUE) / sqrt(n())
    ) %>%
    mutate(
      formatted = sprintf("%.3f±%.3f", mean, se)
    )
  
  # Extract values for KEGG and CAZyme
  kegg_value <- ifelse("KEGG" %in% summary_df$database,
                      summary_df$formatted[summary_df$database == "KEGG"], 
                      NA_character_)
  
  cazyme_value <- ifelse("CAZyme" %in% summary_df$database,
                        summary_df$formatted[summary_df$database == "CAZyme"],
                        NA_character_)
  
  # Return as a data frame row
  data.frame(
    phylum = taxa,
    Levenes_p = round(levene_result, 4),
    Kruskal_p_raw = kruskal_result,  # Keep raw for adjustment
    Kruskal_p = round(kruskal_result, 4),
    KEGG = kegg_value,
    CAZyme = cazyme_value,
    stringsAsFactors = FALSE
  )
}

# Apply function to all OTUs
all_results <- data.frame()

# Get unique OTUs
unique_otus <- unique(df_full_long$OTU)

# Loop through each OTU
all_results <- do.call(rbind, lapply(unique_otus, function(otu) {
  analyze_otu(otu, df_full_long)
}))

# ============================================
# STEP 1: FILTER TO KEEP ONLY SPECIFIED PHYLA
# ============================================
keep <- c("Acidobacteriota", "Actinomycetota", "Bacillota", "Bacteria", 
          "Bacteroidota", "Candidatus Levyibacteriota", "Candidatus Methylomirabilota", 
          "Candidatus Rokuibacteriota", "Candidatus Sysuimicrobiota", "Chloroflexota", 
          "Cyanobacteriota", "Gemmatimonadota", "Methanobacteriota", "Myxococcota", 
          "Nitrososphaerota", "Nitrospirota", "Planctomycetota", "Pseudomonadota", 
          "Thermodesulfobacteriota", "Thermomicrobiota", "Verrucomicrobiota")

# Filter results
filtered_results <- subset(all_results, phylum %in% keep)

# ============================================
# STEP 2: APPLY BH ADJUSTMENT TO FILTERED RESULTS
# ============================================
filtered_results <- filtered_results %>%
  mutate(
    # BH adjustment (only applied to the filtered phyla)
    Kruskal_p_BH = p.adjust(Kruskal_p_raw, method = "BH"),
    # Round the adjusted p-values for display
    Kruskal_p_BH_rounded = round(Kruskal_p_BH, 4),
    # Create significance columns
    significant_raw = ifelse(Kruskal_p_raw < 0.05, "Yes", "No"),
    significant_BH = ifelse(Kruskal_p_BH < 0.05, "Yes", "No")
  )

# ============================================
# STEP 3: VIEW AND SUMMARIZE RESULTS
# ============================================
# View results with adjustments
print(filtered_results %>%
        dplyr::select(phylum, Kruskal_p, Kruskal_p_BH_rounded, 
               significant_raw, significant_BH, Levenes_p, KEGG, CAZyme))



```




```{r}
rm(list = ls())
```


