---
title: "Untitled"
author: "Laura_Hakkinen"
date: "2026-16-05"
output: html_document
---




# Contigs-based analysis of KEGG genes


# 1. Load KEGG data


```{r}
library('phyloseq')
library("cowplot")
library("dplyr")
library("ggplot2")
library("vegan")
library("microbiome")
library("tibble")
library("readxl")
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)

setwd('C:\\Users\\lehakkin\\RStudio_computer')

# load the coverage data
load(file = 'KOfam_metabolism_ps_COVERAGE_FINAL_PNSCO2')
ps

meta <- meta(ps)

ps_RA <- transform(ps, "compositional")
```






# 2. PCoA for KEGG composition



To visualize beta diversity, I will do a PCoA which is metric instead of e.g. non-metric NMDS

I will be following somewhat this tutorial:

[Joey Bernhardt](https://github.com/JoeyBernhardt/NumericalEcology/blob/master/Num._Ecology_Chap._9.md)




```{r}
OTU = as(otu_table(ps_RA), "matrix")
# transpose if necessary
if(taxa_are_rows(ps_RA)){OTU <- t(OTU)}
# Coerce to data.frame
OTU = as.data.frame(OTU)
OTU <- as.matrix(OTU)

bray_dist <- vegan::vegdist(OTU, method="bray")
str(bray_dist)

# use k = 3 so we'll get calculations for three axes
pcoa <- cmdscale(bray_dist, eig=TRUE, k = 3)
ordiplot(scores(pcoa),display="sites", type="points")
```


Let's first make PCoA ordination with axes 1 and 2, and later for 1 and 3.

Env. variables

```{r}
# Post-Hoc Projections of environmental variables
# envfit in vegan projects points onto vectors that have maximum correlation with corresponding environmental variables
pcoa.env12 <- envfit(pcoa, meta[ ,c("pH_H2O", "EC_uScm", "C_g_per_kg", "N_gkg", "TP_gkg", "sample_type", "depth_numerical", "DOC_mgkg", "Pinorg_mgkg", "Porg_mgkg", "log_root", "C_per_N", "Feox_mmolkg", "Alox_mmolkg", "oxides_mmolkg", "PH2O_mgkg")], na.rm = TRUE, choices=c(1:2), permutations = 999)

    

# main effects
pcoa.env12

ef12.adj <- pcoa.env12
pvals.adj <- p.adjust (pcoa.env12$vectors$pvals, method = 'bonferroni')
ef12.adj$vectors$pvals <- pvals.adj
ef12.adj

```


To plot (classical) mds (which is equivalent to PCoA) with ggplot a new datasheet needs to be created which contains the x,y points for each site. You can do this by calling the scores of you mds.




```{r}
site.scrs <- as.data.frame(scores(pcoa, display = "sites")) #save pcoa results into dataframe
site.scrs <- cbind(site.scrs, soil_type = meta$sample_type) #add grouping variable "soil_type" to dataframe
site.scrs <- cbind(site.scrs, depth = meta$depth) #add grouping variable of depth

head(site.scrs)
```


Categorial and continuous variables

```{r}

# first categorial variables
env.scores_cat12 <- as.data.frame(scores(pcoa.env12, display = "factors"))*0.25 # multiply the envfit scores to keep them in the frame of the ordination
env.scores_cat12 <- cbind(env.scores_cat12, env.variables = rownames(env.scores_cat12)) #and then gives them their names

env.scores_cat12 <- cbind(env.scores_cat12, pval = pcoa.env12$factors$pvals) # add pvalues to dataframe
sig.env.scores_cat12 <- subset(env.scores_cat12, pval<=0.05) #subset data to show variables significant at 0.05
sig.env.scores_cat12
# all were significant


# then continous variables
env.scores_cont12 <- as.data.frame(scores(pcoa.env12, display = "vectors"))*0.25 # multiply the envfit scores to keep them in the frame of the ordination
env.scores_cont12 <- cbind(env.scores_cont12, env.variables = rownames(env.scores_cont12)) #and then gives them their names

env.scores_cont12 <- cbind(env.scores_cont12, pval = pcoa.env12$vectors$pvals) # add pvalues to dataframe
sig.env.scores_cont12 <- subset(env.scores_cont12, pval<=0.05) #subset data to show variables significant at 0.05
sig.env.scores_cont12
# all were significant
```



Now we have the relevant information for plotting the ordination in ggplot


```{r}
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")

site.scrs$new_depth <- site.scrs$depth

site.scrs$new_depth <- gsub("...", "-", site.scrs$new_depth, fixed = TRUE)

site.scrs$new_depth[site.scrs$new_depth=="40-"] <- "40-80"

site.scrs$new_depth[site.scrs$new_depth=="0-10"] <- "0-10 cm"
site.scrs$new_depth[site.scrs$new_depth=="10-20"] <- "10-20 cm"
site.scrs$new_depth[site.scrs$new_depth=="20-30"] <- "20-30 cm"
site.scrs$new_depth[site.scrs$new_depth=="30-40"] <- "30-40 cm"
site.scrs$new_depth[site.scrs$new_depth=="40-80"] <- "40-80 cm"
```


Get axis %

```{r}
#First lets run the PCoA with phyloseq to get the axis percentages (plot_ordination gives axis percentages)

GP.ord <- ordinate(ps_RA, "PCoA", "bray")
p2 = plot_ordination(ps_RA, GP.ord, type="samples", color="sample_type", shape="depth")
p2
```

Remember to change the axis percentages accordingly below!!!

## 2.1. Plot the PCoA


```{r}
pcoa.plot <- ggplot() + geom_point(data=site.scrs, aes(Dim1, Dim2, colour = factor(site.scrs$soil_type), shape = factor(site.scrs$new_depth)), size = 6, alpha = 0.6, stroke = 1.5) + theme_cowplot() + theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid")) + labs(colour = "", shape = "") + theme(legend.text = element_text(size = 12), axis.text = element_text(size = 16)) + scale_colour_manual(values = MyPalette) + labs(y = "PCoA2 (22.2%)", x = "PCoA1 (32.1%)")


pcoa.plot

```
Add env. variables to the PCoA

First, simplify the names



```{r}
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "pH_H2O"] <- "pH"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "C_g_per_kg"] <- "C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "N_gkg"] <- "N"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "TP_gkg"] <- "P-tot"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "depth_numerical"] <- "Depth"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "DOC_mgkg"] <- "DOC"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Porg_mgkg"] <- "P-org"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "log_root"] <- "log root"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "C_per_N"] <- "C/N"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Feox_mmolkg"] <- "Fe-ox"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Alox_mmolkg"] <- "Al-ox"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "depth"] <- "Depth"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "log root"] <- "Root biomass"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "S_mg_per_kg"] <- "S"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "Mn_mg_per_kg"] <- "Mn"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "PH2O_mgkg"] <- "P-H2O"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "POM_C_percent"] <- "POM-C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "MAOM_C_percent"] <- "MAOM-C"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "oxides_mmolkg"] <- "Oxides"
sig.env.scores_cont12$env.variables[sig.env.scores_cont12$env.variables == "EC_uScm"] <- "EC"
```

...then plot

```{r}
p1 <- pcoa.plot +
  geom_segment(aes(x = 0, y = 0, xend = Dim1, yend = Dim2), 
               data = sig.env.scores_cont12, 
               size = 1,  # Increased from 1 to 2
               alpha = 0.3, 
               colour = "grey30") + 
  ggrepel::geom_text_repel(data = sig.env.scores_cont12, 
                           aes(x = Dim1, y = Dim2), 
                           colour = "blue", 
                           fontface = "bold", 
                           label = sig.env.scores_cont12$env.variables, 
                           segment.size = 0.15, 
                           box.padding = unit(0.15, "lines"), 
                           point.padding = (0.15), 
                           force = 1, 
                           max.time = 1000, 
                           nudge_y = 0.00, 
                           nudge_x = 0.00)

p1
```





# 3. PERMANOVA of KEGG genes




Calculate Bray-Curtis (dis)similarities

```{r}
ps_RA_bray <- phyloseq::distance(ps_RA, method = "bray")
```


Homogenity of variances

Check that variance homogeneity assumptions hold (to ensure the reliability of the results). If  groups have signif. different spreads the permanova result may be potentially explained by that, rtaher than the groups.

Betadisper first calculates the average distance of group members to the group centroid in multivariate space (generated by a distance matrix). Then, an ANOVA is done to test if the dispersions (variances) of groups are different.

for management type

```{r}
anova(betadisper(ps_RA_bray, meta$sample_type))
```
         

We see that the ANOVA p-value is NOT significant meaning that the **homogeneity of variance assumption is met**


for depth

```{r}
anova(betadisper(ps_RA_bray, meta$depth))
```

We see that the ANOVA p-value is significant meaning that **homogeneity of variance assumption is NOT met**


post hoc

I'll do post hoc analysis with Tukey's test to see which groups differ in relation to their variances

```{r}
TukeyHSD(betadisper(ps_RA_bray, meta$depth))
```

Only the soil layer 40-80 cm differs from 30-40 cm


## 3.1. Initial PERMANOVAs


With forest

First, I will do PERMANOVA so that I include all management types (later without forest)


```{r}
# first with just soil type and strata option
adonis2(formula = ps_RA_bray~ sample_type, data = meta, permutations = 999, method = "bray", by = "margin", strata = meta$depth)
```

Significant

Without forest

```{r}
# subset samples
ps_x <- subset_samples(ps_RA, sample_type != "forest")
meta_subset <- meta(ps_x)


# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_x),
              MARGIN = ifelse(taxa_are_rows(ps_x), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_x = prune_taxa((prev0 > 0), ps_x)
ps_x

b <- phyloseq::distance(ps_x, method = "bray")


set.seed(777)

final <- adonis2(formula = b ~ sample_type, data = meta_subset, permutations = 9999, method = "bray", by = "margin", strata = meta_subset$depth)
final

```


Significant



Depth

```{r}
# then with just depth and strata option
adonis2(formula = ps_RA_bray~ depth, data = meta, permutations = 9999, method = "bray", by = "margin", strata = meta$sample_type)
```

Depth has a larger effect. So, let's put it first in the model


## 3.2. Final PERMANOVA

For the full model it matters which "by" option we choose. When by="terms" significance for each term is calculated sequentially from first to last, so that the order of terms matter. We will use this, because with sequential analysis we will get R2 values that sum up to 1, and will also get the significance and R2 values for each interaction term separately rather than for the interaction alone.

Note! Setting by = "terms" will assess significance for each term (sequentially from first to last), by = "margin" will assess the marginal effects of the terms (each marginal term analysed in a model with all other variables)




```{r}
set.seed(270285)
final <- adonis2(formula = ps_RA_bray ~ depth * sample_type, data = meta, permutations = 9999, method = "bray", by = "terms")
final

```




## 3.3. Add PERMANOVA results to PCoA

```{r}
library(ggpubr) ## for getting stat_regline_equation
library("ggpmisc")

KEGG_PCoA <- p1 + 
  annotate(
    geom = "text",
    x = -0.20,
    y = 0.35,
    label = "PERMANOVA \nSoil layer:Treatment: R²=0.09; p=0.044 \nSoil layer: R²=0.27; p<0.001 \nTreatment: R²=0.06; p<0.001 \nModel R²=0.42",
    hjust = 0,
    size = 4  # Increase this value until text matches your desired size (default is 3.88)
  )

KEGG_PCoA


```





# 4. Heatmaps of KEGG

## 4.1. CO2 fixation

### 4.1.1. pathway



I will do heatmap for RA values.

For the final heatmap figure the RA values will be standardizes (z-transformed)

Note! I will not do Z-transforming the data before plotting the heatmap. Instead I will do row scaling in the Pheatmap stage, which means that each gene's relative abundance is standardized



Subset the CO2-fixing


```{r}
ps_RA <- transform(ps, "compositional")
ps_co2fix <- subset_taxa(ps_RA, metabolism == "CO2_fixing")
ps_co2fix

ps_co2fix <- aggregate_rare(ps_co2fix, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_co2fix

```



```{r}
meta <- meta(ps_co2fix)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_co2fix_z <- microbiome::transform(ps_co2fix, 'Z', log10 = FALSE)
ps_co2fix_z <- ps_co2fix

data_subset <- as(otu_table(ps_co2fix_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 7)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```





```{r}
# 1. Prepare group data
group <- as.data.frame(tax_table(ps))
group <- subset(group, metabolism=="CO2_fixing")
group <- group[!duplicated(group[c('pathway')]), ]
rownames(group) <- group$pathway  # Ensure pathway names are row names
group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```


```{r}
row.names(f4_ordered)
```


```{r}
# Change pathway names and remove unwanted

## these will be removed: C4_Dicarboxylic_acid_cycle
# row of data frame


row.names(f4_ordered)[1] <- "Arnon-Buchanan cycle"  
row.names(f4_ordered)[2] <- "Calvin cycle"
row.names(f4_ordered)[3] <- "Hydroxypropionate bicycle"  
row.names(f4_ordered)[4] <- "Wood-Ljungdahl pathway" 
```




```{r}
library("ggplotify")
library("pheatmap")
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = TRUE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, gaps_row = row_gaps, annotation_names_col = FALSE, scale = "row"))

p2
```



### 4.2.2. gene



```{r}
ps_RA <- transform(ps, "compositional")

ps_co2fix <- subset_taxa(ps_RA, metabolism == "CO2_fixing")
ps_co2fix

ps_co2fix <- aggregate_rare(ps_co2fix, level = 'gene', detection = 0/100, prevalence = 0/140)
ps_co2fix

```




```{r}
meta <- meta(ps_co2fix)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_co2fix_z <- microbiome::transform(ps_co2fix, 'Z', log10 = FALSE)
ps_co2fix_z <- ps_co2fix

data_subset <- as(otu_table(ps_co2fix_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 13)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```



```{r}
# 1. Prepare group data
group <- as.data.frame(tax_table(ps))
group <- subset(group, metabolism=="CO2_fixing")
# remove duplicate genes
group <- group[!duplicated(group[c('gene')]), ]

group$gene[group$metabolism == "CO2_fixing" & !is.na(group$metabolism)]
rownames(group) <- group$gene  # Ensure gene names are row names
group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```



```{r}
library("ggplotify")
library("pheatmap")
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, gaps_row = row_gaps, annotation_names_col = FALSE, scale = "row"))

p2
```


## 4.2. P-cycling


### 4.2.1. pathway

```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- subset_taxa(ps_RA, metabolism=="P_cycling")
ps_x

x <- as.data.frame(tax_table(ps_x))
x <- x[, c(6, 1:5)]

tax_table(ps_x) <- tax_table(as.matrix(x))

ps_p <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_p
```


```{r}
meta <- meta(ps_p)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_p_z <- microbiome::transform(ps_p, 'Z', log10 = FALSE)
ps_p_z <- ps_p

data_subset <- as(otu_table(ps_p_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 6)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```



```{r}
# 1. Prepare group data
group <- as.data.frame(tax_table(ps_p))
#group <- subset(group, metabolism == "P_cycling")
#rownames(group) <- group$gene  # Ensure gene names are row names
#group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```


```{r}
row.names(f4_ordered)
```



```{r}
row.names(f4_ordered)[1] <- "P starvation"  
row.names(f4_ordered)[2] <- "P-inorganic solubilization"
row.names(f4_ordered)[3] <- "P-organic mineralization"  
```




```{r}
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, annotation_names_col = FALSE, scale = "row"))
#, gaps_row = row_gaps
p2
```

### 4.2.2. gene


```{r}
ps_RA <- transform(ps, "compositional")
ps_p <- subset_taxa(ps_RA, metabolism=="P_cycling")
ps_p

ps_p <- aggregate_rare(ps_p, level = 'gene', detection = 0/100, prevalence = 0/140)
ps_p
```


```{r}
meta <- meta(ps_p)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_p_z <- microbiome::transform(ps_p, 'Z', log10 = FALSE)
ps_p_z <- ps_p

data_subset <- as(otu_table(ps_p_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 24)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```



```{r}
# 1. Prepare group data

group <- as.data.frame(tax_table(ps_RA))
group <- subset(group, metabolism == "P_cycling")
rownames(group) <- group$gene  # Ensure gene names are row names
group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```





```{r}
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, gaps_row = row_gaps, annotation_names_col = FALSE, scale = "row"))

p2
```




## 4.3. N-cycling



### 4.3.1. pathway



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- subset_taxa(ps_RA, metabolism=="N_cycling")
ps_x

x <- as.data.frame(tax_table(ps_x))
x <- x[, c(6, 1:5)]

tax_table(ps_x) <- tax_table(as.matrix(x))

ps_p <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_p
```


```{r}
meta <- meta(ps_p)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_p_z <- microbiome::transform(ps_p, 'Z', log10 = FALSE)
ps_p_z <- ps_p

data_subset <- as(otu_table(ps_p_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 9)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```




```{r}
# 1. Prepare group data
group <- as.data.frame(tax_table(ps_p))
#group <- subset(group, metabolism == "P_cycling")
#rownames(group) <- group$gene  # Ensure gene names are row names
#group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

#row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```


```{r}
# Create a copy of the data frame to avoid modifying the original
f4_result <- f4_ordered

# Add denitrify_DNRA values to both denitrify and DNRA rows
f4_result["denitrify", ] <- f4_result["denitrify", ] + f4_result["denitrify_DNRA", ]
f4_result["DNRA", ] <- f4_result["DNRA", ] + f4_result["denitrify_DNRA", ]

# Remove the denitrify_DNRA row
f4_result <- f4_result[rownames(f4_result) != "denitrify_DNRA", ]

f4_ordered <- f4_result
```


```{r}
row.names(f4_ordered)
```



```{r}
row.names(f4_ordered)[1] <- "ANRA"  
row.names(f4_ordered)[2] <- "Denitrification"
row.names(f4_ordered)[3] <- "DNRA"  
row.names(f4_ordered)[4] <- "N fixation" 
row.names(f4_ordered)[5] <- "Nitrification" 

f4_ordered <- f4_ordered[c(4, 5, 2, 1, 3),]
```



```{r}
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = TRUE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, annotation_names_col = FALSE, scale = "row"))

#, gaps_row = row_gaps

p2
```



### 4.3.2. gene


Subset the N-cycling


```{r}
ps_RA <- transform(ps, "compositional")
ps_n <- subset_taxa(ps_RA, metabolism=="N_cycling")
ps_n

ps_n <- aggregate_rare(ps_n, level = 'gene', detection = 0/100, prevalence = 0/140)
ps_n
```




```{r}
meta <- meta(ps_n)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_n_z <- microbiome::transform(ps_n, 'Z', log10 = FALSE)
ps_n_z <- ps_n

data_subset <- as(otu_table(ps_n_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 29)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```



```{r}
# 1. Prepare group data

group <- as.data.frame(tax_table(ps_RA))
group <- subset(group, metabolism == "N_cycling")
rownames(group) <- group$gene  # Ensure gene names are row names
group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```




```{r}
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, gaps_row = row_gaps, annotation_names_col = FALSE, scale = "row"))

p2
```





## 4.4. S-cycling



### 4.4.1. pathway



```{r}
ps_RA <- transform(ps, "compositional")

ps_x <- subset_taxa(ps_RA, metabolism=="S_cycling")
ps_x

x <- as.data.frame(tax_table(ps_x))
x <- x[, c(6, 1:5)]

tax_table(ps_x) <- tax_table(as.matrix(x))

ps_p <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_p
```


```{r}
meta <- meta(ps_p)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_p_z <- microbiome::transform(ps_p, 'Z', log10 = FALSE)
ps_p_z <- ps_p

data_subset <- as(otu_table(ps_p_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 9)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```







```{r}
# 1. Prepare group data
group <- as.data.frame(tax_table(ps_p))
#group <- subset(group, metabolism == "P_cycling")
#rownames(group) <- group$gene  # Ensure gene names are row names
#group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```


```{r}
row.names(f4_ordered)
```

```{r}
# Create a copy of the data frame to avoid modifying the original
f4_result <- f4_ordered

# Add denitrify_DNRA values to both denitrify and DNRA rows
f4_result["ASR", ] <- f4_result["ASR", ] + f4_result["DSR_ASR", ]
f4_result["DSR", ] <- f4_result["DSR", ] + f4_result["DSR_ASR", ]

# Remove the denitrify_DNRA row
f4_result <- f4_result[rownames(f4_result) != "DSR_ASR", ]

f4_ordered <- f4_result
```


```{r}
row.names(f4_ordered)
```



```{r}
row.names(f4_ordered)[1] <- "ASR"  
row.names(f4_ordered)[2] <- "DSR"
row.names(f4_ordered)[3] <- "S oxidation"  
row.names(f4_ordered)[4] <- "S reduction" 
row.names(f4_ordered)[5] <- "Sulfide oxidation" 

f4_ordered <- f4_ordered[c(5, 3, 4, 1, 2),]
```




```{r}
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, annotation_names_col = FALSE, scale = "row"))
#, gaps_row = row_gaps
p2
```


### 4.4.2. gene


Subset the S-cycling

```{r}
ps_RA <- transform(ps, "compositional")
ps_n <- subset_taxa(ps_RA, metabolism=="S_cycling")
ps_n

ps_n <- aggregate_rare(ps_n, level = 'gene', detection = 0/100, prevalence = 0/140)
ps_n
```




```{r}
meta <- meta(ps_n)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"

meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))

# Z-transform
#ps_n_z <- microbiome::transform(ps_n, 'Z', log10 = FALSE)
ps_n_z <- ps_n

data_subset <- as(otu_table(ps_n_z), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 29)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```



```{r}
# 1. Prepare group data

group <- as.data.frame(tax_table(ps_RA))
group <- subset(group, metabolism == "S_cycling")
rownames(group) <- group$gene  # Ensure gene names are row names
group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$pathway <- factor(group$pathway)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$pathway), ]

row_gaps <- cumsum(table(group$pathway))[-length(table(group$pathway))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```




```{r}
p2 <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, gaps_row = row_gaps, annotation_names_col = FALSE, scale = "row"))

p2
```







# 5. Differential abundance analysis: MaAsLin2



MaAsLin2 (Microbiome Multivariate Analysis)
Handles relative abundance data.

Works with generalized linear models (GLMs).


I will be using NPS-cycling genes only (and one unknown which can be ignored)


in the final model i want to use relative abundances and use CPLM instead of the LM model. This is because with LM model I would need to use LOG transformation or something similar, but this is problem with zero values. I also dont want to add a pseudocount as this creates other issues. In CPLM I can exclude the LOG transformation. In fact "All the non-LM models use an intrinsic log link transformation due to their close connection to GLMs and they are recommended to be run with transform = NONE."

The CPLM is meant to be used with non-count data as RA data is.




## 5.1. Metabolic level

Forest excluded due to too small replicate number!



```{r}
ps_x <- aggregate_rare(ps_RA, level = 'metabolism', detection = 0/100, prevalence = 0/140)
ps_x

# remove "Unknown"
ps_x <- subset_taxa(ps_x, metabolism!="Unknown")
ps_x

tax <- as.data.frame(tax_table(ps_x))

```

```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_x, sample_type!="forest")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\Metabolic_level_WITH_CO2')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

# this is how the analysis was run

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "maaslin_METABOLISM_TREATMENT_and_LAYER",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )
```






## 5.2. CO2 fixing: genes


Forest excluded due to too small replicate number!

```{r}
ps_x <- subset_taxa(ps_RA, metabolism=="CO2_fixing")
ps_x

```



```{r}
ps_RA_phyla_aggr_good <- microbiome::aggregate_rare(ps_x, level = 'gene', detection = 0/140, prevalence = 0/140)
ps_RA_phyla_aggr_good
```



### 5.2.1. With Meadow

```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_phyla_aggr_good, sample_type!="forest")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CO2_fixing_GENE')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
# input_data = gene_data,
#  input_metadata = metadata,
#  output = "TREATMENT_and_LAYER",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )
```





### 5.2.2. Without Meadow




```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_phyla_aggr_good, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CO2_fixing_GENE')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "TREATMENT_and_LAYER_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional'
#  )
```





## 5.3. CO2 fixing: pathway


Forest excluded due to too small replicate number!

```{r}
ps_x <- subset_taxa(ps_RA, metabolism=="CO2_fixing")
ps_x
```




```{r}
ps_RA_phyla_aggr_good <- microbiome::aggregate_rare(ps_x, level = 'pathway', detection = 0/140, prevalence = 0/140)
ps_RA_phyla_aggr_good
```


### 5.3.1. With meadow



```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_phyla_aggr_good, sample_type!="forest")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```




```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CO2_fixing_pathway')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "TREATMENT_and_LAYER_WITH_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )

```




### 5.3.2. Without meadow



```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_phyla_aggr_good, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```




```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\CO2_fixing_pathway')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "TREATMENT_and_LAYER_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional'
#  )

```






## 5.4. P cycling: genes



Forest excluded due to too small replicate number!

```{r}
ps_x <- subset_taxa(ps_RA, metabolism=="P_cycling")
ps_x
```




```{r}
ps_RA_phyla_aggr_good <- microbiome::aggregate_rare(ps_x, level = 'gene', detection = 0/140, prevalence = 0/140)
ps_RA_phyla_aggr_good
```




### 5.4.1. With Meadow

```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_phyla_aggr_good, sample_type!="forest")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\P_cycling_GENE')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "TREATMENT_and_LAYER",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Meadow'
#  )
```





### 5.4.2. Without Meadow




```{r}
#To remove those from your phyloseq object
ps_y <- subset_samples(ps_RA_phyla_aggr_good, sample_type!="forest" & sample_type!="meadow")
#This will return a ps object with the samples removed
ps_y

# You have only submitted samples, not OTUs. Define prevalence of each taxa
# (in how many samples did each taxa appear at least once)
prev0 = apply(X = otu_table(ps_y),
              MARGIN = ifelse(taxa_are_rows(ps_y), yes = 1, no = 2),
              FUN = function(x){sum(x > 0)})

# Execute prevalence filter, using `prune_taxa()` function
ps_y = prune_taxa((prev0 > 0), ps_y)
ps_y

```

```{r}
# Run MaAsLin2
library(Maaslin2)

setwd('C:\\Users\\lehakkin\\RStudio_computer\\GENE_MaAsLin2\\P_cycling_GENE')

## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

#fit_data <- Maaslin2(
#  input_data = gene_data,
#  input_metadata = metadata,
#  output = "TREATMENT_and_LAYER_WO_MEADOW",
#  fixed_effects = c('Layer', 'Treatment'), # my grouping variable(s)
#  normalization = "NONE",          # Use CLR for relative abundances
#  transform = "NONE",             # No extra transform (data is already proportional)
#  analysis_method = "CPLM", #DEFAULT:
#  correction = "BH",               # Benjamini-Hochberg FDR correction
#  plot_heatmap = TRUE,
#  reference = 'Treatment,Conventional'
#  )
```



## 5.5. P-cycling: pathway

etc for different metabolism:

N-cycling

S-cycling





# 6. METABOLISM FIGURES


Metabolic pathways



```{r}
meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_RA) <- sample_data(meta)
ps_x <- ps_RA
ps_x
```



```{r}
ps_y <- aggregate_rare(ps_x, level = 'metabolism', detection = 0/100, prevalence = 0/140)
ps_y

ps_y <- subset_taxa(ps_y, metabolism!="CO2_fixing")
ps_y
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```



```{r}
names(gene_data)
```




## N, P, S combined 



```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


Reshape df

```{r}

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)


df_long <- df %>%
  pivot_longer(
    cols = c(N_cycling, P_cycling, S_cycling),  # Columns to combine
    names_to = "Metabolism",  # New column for GH/GT
    values_to = "Proportion"  # New column for values
  )
```



```{r}
fig <- df_long %>%
  group_by(sample_type, depth_numerical, Metabolism) %>%
  summarise(
    mean = mean(Proportion, na.rm = TRUE),
    se = sd(Proportion, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  facet_wrap(~ Metabolism, 
             scales = "free_x",
             labeller = labeller(Metabolism = c(
               "N_cycling" = "N cycling genes",
               "P_cycling" = "P cycling genes", 
               "S_cycling" = "S cycling genes"
             ))) +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "Proportion",
    color = "Treatment"
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 22, face = "bold"),
    legend.text = element_text(size = 22),
    legend.title = element_text(size = 22),
    plot.title = element_text(size = 22, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    strip.text = element_text(size = 22, face = "bold")
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) + theme(legend.title=element_blank())

print(fig)
```







## 6.1. N_cycling




```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

N_cycling <- ggplot(df, aes(x = Layer, y = N_cycling, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "N cycle") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

N_cycling

```




```{r}
names(gene_data)
```



## 6.2. P_cycling



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

P_cycling <- ggplot(df, aes(x = Layer, y = P_cycling, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "P cycle") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

P_cycling

```



```{r}
names(gene_data)
```



## 6.3. S_cycling




```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

S_cycling <- ggplot(df, aes(x = Layer, y = S_cycling, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "S cycle") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

S_cycling

```








## 6.4. CO2 fixing pathway




```{r}
ps_y <- aggregate_rare(ps_x, level = 'metabolism', detection = 0/100, prevalence = 0/140)
ps_y

ps_y <- subset_taxa(ps_y, metabolism=="CO2_fixing")
ps_y
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```





```{r}
# create your own color palette for sample types
MyPalette <- c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c")
```


Reshape df

```{r}

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)


df_long <- df %>%
  pivot_longer(
    cols = c(CO2_fixing),  # Columns to combine
    names_to = "Metabolism",  # New column for GH/GT
    values_to = "Proportion"  # New column for values
  )
```



```{r}
fig <- df_long %>%
  group_by(sample_type, depth_numerical, Metabolism) %>%
  summarise(
    mean = mean(Proportion, na.rm = TRUE),
    se = sd(Proportion, na.rm = TRUE) / sqrt(n())
  ) %>%
  ungroup() %>%
  ggplot(aes(
    y = mean,
    x = depth_numerical,
    color = sample_type
  )) +
  geom_line(position = position_dodge(1.2)) +
  geom_point(size = 2, position = position_dodge(1.2)) +
  geom_errorbar(
    aes(ymin = mean - se, ymax = mean + se),
    width = 0.3,
    position = position_dodge(1.2),
    linetype = "solid",
    alpha = 0.5
  ) +
  coord_flip() +
  scale_x_reverse() +
  facet_wrap(~ Metabolism, 
             scales = "free_x",
             labeller = labeller(Metabolism = c(
               "CO2_fixing" = "C fixation key genes"
             ))) +
  scale_colour_manual(values = MyPalette) +
  labs(
    x = "Soil layer",
    y = "Proportion",
    color = "Treatment"
  ) +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 22, face = "bold"),
    legend.text = element_text(size = 22),
    legend.title = element_text(size = 22),
    plot.title = element_text(size = 22, hjust = 0.5),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    strip.text = element_text(size = 22, face = "bold")
  ) +
  scale_x_reverse(
    breaks = c(60, 35, 25, 15, 5),
    labels = c("40-80 cm", "30-40 cm", "20-30 cm", "10-20 cm", "0-10 cm")
  ) + theme(legend.title=element_blank())

print(fig)
```





Soil layer Treatment



```{r}
meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_RA) <- sample_data(meta)
ps_x <- subset_taxa(ps_RA, metabolism=="CO2_fixing")
ps_x
```



```{r}
ps_y <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_y
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```




 Calvin_cycle





```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Calvin_cycle <- ggplot(df, aes(x = Layer, y = Calvin_cycle, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Calvin cycle\n") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Calvin_cycle

```



```{r}
names(gene_data)
```




```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

hydroxypropionate_bicycle <- ggplot(df, aes(x = Layer, y = hydroxypropionate_bicycle, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Hydroxy-propionate-\nbi-cycle") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

hydroxypropionate_bicycle

```




```{r}
names(gene_data)
```


Arnon_Buchanan_cycle



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Arnon_Buchanan_cycle <- ggplot(df, aes(x = Layer, y = Arnon_Buchanan_cycle, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Arnon-Buchanan-\ncycle") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Arnon_Buchanan_cycle

```



```{r}
names(gene_data)
```




Wood_Ljungdahl_pathway


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Wood_Ljungdahl_pathway <- ggplot(df, aes(x = Layer, y = Wood_Ljungdahl_pathway, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Wood-Ljungdahl-\npathway") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Wood_Ljungdahl_pathway

```







## 6.5. P-cycling pathway




```{r}
meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_RA) <- sample_data(meta)
ps_x <- subset_taxa(ps_RA, metabolism=="P_cycling")
ps_x

x <- as.data.frame(tax_table(ps_x))
x <- x[, c(1, 5, 6, 2, 4, 3)]
x <- as.matrix(x)
tax_table(ps_x) <- tax_table(x)
```



```{r}
ps_y <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_y
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```




P_starv



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

P_starv <- ggplot(df, aes(x = Layer, y = P_starv, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "P starvation\n") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

P_starv

```



```{r}
names(gene_data)
```




Pinorg_solub


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Pinorg_solub <- ggplot(df, aes(x = Layer, y = Pinorg_solub, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "P-inorganic\nsolubilization") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Pinorg_solub

```




```{r}
names(gene_data)
```




Porg_miner



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Porg_miner <- ggplot(df, aes(x = Layer, y = Porg_miner, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "P-organic\nmineralization") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Porg_miner

```



```{r}
names(gene_data)
```


## 6.6. N-cycling pathway






```{r}
meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_RA) <- sample_data(meta)
ps_x <- subset_taxa(ps_RA, metabolism=="N_cycling")
ps_x

x <- as.data.frame(tax_table(ps_x))
x <- x[, c(1, 5, 6, 2, 4, 3)]
x <- as.matrix(x)
tax_table(ps_x) <- tax_table(x)
```



```{r}
ps_y <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_y
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```



First, lets add the denitrify_DNRA to both denitrify and DNRA

```{r}
# Add denitrify_DNRA to denitrify
gene_data$denitrify <- gene_data$denitrify + gene_data$denitrify_DNRA

# Add denitrify_DNRA to DNRA
gene_data$DNRA <- gene_data$DNRA + gene_data$denitrify_DNRA
```





ANRA


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

ANRA <- ggplot(df, aes(x = Layer, y = ANRA, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "ANRA\n") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

ANRA

```



```{r}
names(gene_data)
```




denitrify


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

denitrify <- ggplot(df, aes(x = Layer, y = denitrify, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Denitrification") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

denitrify

```




```{r}
names(gene_data)
```




DNRA



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

DNRA <- ggplot(df, aes(x = Layer, y = DNRA, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "DNRA") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

DNRA

```


```{r}
names(gene_data)
```




N_fix



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

N_fix <- ggplot(df, aes(x = Layer, y = N_fix, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "N fixation") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

N_fix

```




```{r}
names(gene_data)
```



nitrify



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

nitrify <- ggplot(df, aes(x = Layer, y = nitrify, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Nitrification") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

nitrify

```




```{r}
names(gene_data)
```




## 6.7. S-cycling pathway





```{r}
meta$sample_type <- factor(meta$sample_type, levels = c("forest", "meadow", "organic", "conventional"))
sample_data(ps_RA) <- sample_data(meta)
ps_x <- subset_taxa(ps_RA, metabolism=="S_cycling")
ps_x

x <- as.data.frame(tax_table(ps_x))
x <- x[, c(1, 5, 6, 2, 4, 3)]
x <- as.matrix(x)
tax_table(ps_x) <- tax_table(x)
```



```{r}
ps_y <- aggregate_rare(ps_x, level = 'pathway', detection = 0/100, prevalence = 0/140)
ps_y
```



```{r}
## get the otu table
gene_data <- as.data.frame(otu_table(ps_y))
gene_data <- as.data.frame(t(gene_data))

metadata <- meta(ps_y)

metadata$Layer <- gsub("...", "-", metadata$depth, fixed = TRUE)

metadata$Layer[metadata$Layer=="40-"] <- "40-80"


metadata$Layer[metadata$Layer=="0-10"] <- "0-10 cm"
metadata$Layer[metadata$Layer=="10-20"] <- "10-20 cm"
metadata$Layer[metadata$Layer=="20-30"] <- "20-30 cm"
metadata$Layer[metadata$Layer=="30-40"] <- "30-40 cm"
metadata$Layer[metadata$Layer=="40-80"] <- "40-80 cm"

metadata$Layer <- as.factor(metadata$Layer)
levels(metadata$Layer)

metadata$Treatment <- metadata$sample_type
metadata$Treatment <- as.character(metadata$Treatment)

metadata$Treatment[metadata$Treatment=="forest"] <- "Forest"
metadata$Treatment[metadata$Treatment=="meadow"] <- "Meadow"
metadata$Treatment[metadata$Treatment=="organic"] <- "Organic"
metadata$Treatment[metadata$Treatment=="conventional"] <- "Conventional"

```

```{r}
names(gene_data)
```

First, lets add the DSR_ASR to both DSR and ASR

```{r}
# Add denitrify_DNRA to denitrify
gene_data$DSR <- gene_data$DSR + gene_data$DSR_ASR

# Add denitrify_DNRA to DNRA
gene_data$ASR <- gene_data$ASR + gene_data$DSR_ASR
```



ASR


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

ASR <- ggplot(df, aes(x = Layer, y = ASR, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "ASRA") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

ASR

```



```{r}
names(gene_data)
```



DSR


```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

DSR <- ggplot(df, aes(x = Layer, y = DSR, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "DSR") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

DSR

```




```{r}
names(gene_data)
```


S_ox



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

S_ox <- ggplot(df, aes(x = Layer, y = S_ox, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "S oxidation") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

S_ox

```


```{r}
names(gene_data)
```



S_red



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

S_red <- ggplot(df, aes(x = Layer, y = S_red, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "S reduction") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

S_red

```




```{r}
names(gene_data)
```



Sulfide_ox



```{r}
my_colour <-  c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A")

#left join
df <- merge(metadata, gene_data, by=0, all.x=TRUE)

Sulfide_ox <- ggplot(df, aes(x = Layer, y = Sulfide_ox, fill = Layer)) +
  geom_boxplot(outlier.shape = NA) +  
  geom_jitter(
    aes(fill = Layer),
    shape = 21,
    color = "black",
    stroke = 0.3,
    width = 0.2,
    alpha = 0.5,
    size = 2
  ) +  
  scale_fill_manual(values = my_colour) +
  stat_summary(fun = "mean", geom = "point", shape = 15, size = 3, color = "black") +  
  labs(x = "", y = "Sulfide oxidation") +  
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +  # Facet by sample_type
  theme_cowplot() +  
  theme(
    legend.position = "none",  
    axis.title.y = element_text(size = 14),  
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),  # Rotate x-axis labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "gray90"),  # Background for facet labels
    strip.text = element_text(face = "bold", size = 12),  # Style for facet labels
    panel.spacing = unit(1, "cm")
  )

Sulfide_ox

```




```{r}
names(gene_data)
```




# 7. METABOLISM Heatmap





```{r}
ps_x <- aggregate_rare(ps_RA, level = 'metabolism', detection = 0/100, prevalence = 0/140)
ps_x

# remove "Unknown"
ps_x <- subset_taxa(ps_x, metabolism!="Unknown")
ps_x

tax <- as.data.frame(tax_table(ps_x))
```

I will do heatmap for RA values.

For the final heatmap figure the RA values will be standardizes (z-transformed)

Note! I will not do Z-transforming the data before plotting the heatmap. Instead I will do row scaling in the Pheatmap stage, which means that each gene's relative abundance is standardized





```{r}
meta <- meta(ps_x)

meta$Depth <- gsub("...", "-", meta$depth, fixed = TRUE)

meta$Depth[meta$Depth=="40-"] <- "40-80"


meta$Depth[meta$Depth=="0-10"] <- "0-10 cm"
meta$Depth[meta$Depth=="10-20"] <- "10-20 cm"
meta$Depth[meta$Depth=="20-30"] <- "20-30 cm"
meta$Depth[meta$Depth=="30-40"] <- "30-40 cm"
meta$Depth[meta$Depth=="40-80"] <- "40-80 cm"


data_subset <- as(otu_table(ps_x), "matrix")
data_subset <- as.matrix(data_subset)

# add annotations "Depth" and "Treatment"
my_sample_col <- data.frame(meta[c("Depth", "sample_type")], row.names = row.names(meta))
colnames(my_sample_col) <- c("Depth", "Treatment")


x <- as.data.frame(t(data_subset))
y <- as.data.frame(my_sample_col)
colnames(y) <- c("Depth", "sample_type")
z <- dplyr::left_join(rownames_to_column(x), rownames_to_column(y), by=c("rowname" = "rowname"))
colnames(z)[1] <- "ID"

# make into mean per treatment and soil layer
f <- z %>%
  group_by(sample_type, Depth) %>%
  summarise_all(mean)

# remove ID column
f <- f[ , -3]

# make new ID column, which is the soiltypedepth
library(stringr)
f$ID <- str_c(f$sample_type, '_', f$Depth)

# make new df with just sample type, depth and sampletypedepth
df <- f[c("sample_type", "Depth", "ID")]
# remove extra columns from f
f2 <- f[, -c(1, 2, 7)]
# column into rownames
rownames(f2) <- f$ID
# make into numeric matrix
f3 <- data.matrix(f2, rownames.force = NA)
f4 <- t(f3)

# same for df
# remove extra columns from df
df2 <- df[ , -c(3)]
# column into rownames
rownames(df2) <- df$ID
my_sample_col2 <- df2
my_sample_col3 <- as.data.frame(my_sample_col2)
rownames(my_sample_col3) <- df$ID
# lets add annotations of samples

# add annotations "Depth" and "soil_type" and change order
colnames(my_sample_col3) <- c("Treatment", "Depth")
my_sample_col3 <- my_sample_col3[, c(2,1)]

```









```{r}
# 1. Prepare group data
group <- tax
group <- group[match(rownames(f4), rownames(group)), ]  # Match order with f4

# Clean pathway names (no annotation needed)
group$metabolism <- factor(group$metabolism)  # Ensure it's a factor

# Order f4 rows by pathway (no gaps needed if no annotation)
f4_ordered <- f4[order(group$metabolism), ]

row_gaps <- cumsum(table(group$metabolism))[-length(table(group$metabolism))]



#Create color palette
my_colour = list(
  "Treatment" = c(forest = "#1167b1", meadow = "#fbc02d", organic = "#8a8a8a", conventional =  "#b71c1c"),
  Depth = c('0-10 cm' = "#387212", '10-20 cm' = "#ADC476", '20-30 cm' = "#D8D2BA",'30-40 cm' = "#907852", '40-80 cm' = "#6A4C3A"))

# add gap between sample_types
gp_col = which(diff(as.numeric(factor(my_sample_col3$Treatment)))!=0)
```


```{r}
row.names(f4_ordered)
```


```{r}
# Change pathway names
# row of data frame

row.names(f4_ordered)[1] <- "CO2 fixing" 
row.names(f4_ordered)[2] <- "N cycling" 
row.names(f4_ordered)[3] <- "P cycling" 
row.names(f4_ordered)[4] <- "S cycling" 
```






```{r}
library(ggplotify)
library(pheatmap)
p <- as.ggplot(function() pheatmap(f4_ordered, cluster_cols = FALSE, cluster_rows = FALSE, annotation_col = my_sample_col3, annotation_colors = my_colour, gaps_col = gp_col, color=colorRampPalette(c("navy", "white", "red"))(50), show_colnames = FALSE, legend = TRUE, border_color = NA, cellheight = 16, fontsize = 14, fontsize_row = 14, annotation_names_row = FALSE, gaps_row = FALSE, annotation_names_col = FALSE, scale = "row"))

p
```







```{r}
rm(list = ls())
```







